---
title: "TBU"
author: "Zhuo Diao, Nikita Gupta"
date: "**March 30, 2026**"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
always_allow_html: true
urlcolor: blue
editor_options: 
  markdown: 
    wrap: 72
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---


# Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "", prompt = TRUE, include = TRUE)#, cache=TRUE, autodep = TRUE)
knitr::opts_chunk$set(fig.align="center") 
```


```{r}
# Install remotes if you don't have it
if (!require("remotes")) install.packages("remotes")

# Install the micd package from the BIPS repository
remotes::install_github("bips-hb/micd")

#  Load the library
library(micd)
library(pcalg)
```

```{r}

library(tidyverse) #installed tidyverse
library(nlme)
library(ggplot2)
library(here)
library(kableExtra)
library(modelsummary)
library(dplyr)
library(performance)
library(glmnet)
library(pcalg)
library(gridExtra)
library(bnlearn)
library(igraph)
library(gt)
library(gtExtras)
library(corrplot)
library(ggpubr)
library(dagitty)

bootstrap_sample_size <- 1000

set.seed(516)
getwd()
```


```{r load data}
data <- read.csv(here("low2076upd.csv"))
colSums(is.na(data)) #no NAs in any columns
nrow(data) #500 rows
colnames(data)
```

# Data manipulation

```{r factorize, }
#factorize cateogorical variables
data$Y <- as.factor(data$Y)
data$A <- as.factor(data$A)
data$SEX <- as.factor(data$SEX)
data$EDUCATION <- as.factor(data$EDUCATION)
data$PAY_0 <- as.factor(data$PAY_0)
data$PAY_2 <- as.factor(data$PAY_2)
data$PAY_3 <- as.factor(data$PAY_3)
data$PAY_4 <- as.factor(data$PAY_4)
data$PAY_5 <- as.factor(data$PAY_5)
data$PAY_6 <- as.factor(data$PAY_6)
```

```{r create subset of covariates}
#create list of ALL covariates - everything except A and Y
cov_all <- setdiff(names(data), c("Y", "A"))
print(cov_all)
```

```{r create enrolled and not enrolled datasets, untransformed data}
# Create a version of the data where everyone is enrolled, and no one is enrolled, for outcome regression
data_all_enrolled <- data
data_all_enrolled$A <- 1
data_all_enrolled$A <- as.factor(data_all_enrolled$A)

data_no_enrolled <- data
data_no_enrolled$A <- 0
data_no_enrolled$A <- as.factor(data_no_enrolled$A)
print(unique(data_all_enrolled$A)) #looks good
print(unique(data_no_enrolled$A))
```


# EDA


# 1 Now the Analysis, for all variables!



# All Variables: First, untransformed data

# All Variables, Outcome regression

Basically, model the outcome (Y) of defaulting. 

1. Train the model to understand how every variable including PPP enrollment A affects prob of defaulting Y. 

2. Take all my customers (A=0 and A=1), and see what their default rate would be if they had A=1.

3. Take all my customers (A=0 and A=1), and see what their default rate would be if they had A=0.

4. ACE is the difference between the two.

```{r outcome reg all vars}
formula_or_all_var <- as.formula(paste("Y ~ A +", paste(cov_all, collapse = "+"))) #now A is part of this
model_or_all_var <- glm(formula_or_all_var, data = data, family = binomial) #binomial since 0 or 1 default

print("formula")
print(formula_or_all_var)
# Predict default probability for world of everyone enrolled, and no one enrolled
data$prob_Y_if_A1_all_var <- predict(model_or_all_var, newdata = data_all_enrolled, type = "response") #type=response gives 0-1 prob
EY_a1_or_all_var <- mean(data$prob_Y_if_A1_all_var)
data$prob_Y_if_A0_all_var <- predict(model_or_all_var, newdata = data_no_enrolled, type = "response")
EY_a0_or_all_var <- mean(data$prob_Y_if_A0_all_var)

ace_or_all_var <- EY_a1_or_all_var - EY_a0_or_all_var

#print means
print("Default likelihood (Full Set, OR), A=1 then A=0")
print(EY_a1_or_all_var)
print(EY_a0_or_all_var)

print(paste("OR ACE (Full Set):", round(ace_or_all_var,4))) 
```

OR ACE for all vars is 0.2108, meaning enrolling in the PPP is associated with 21% increase in probability of missing a payment. This seems unintuitive...


# All Variables, IPW
Basically, we know that PPP enrollment is not split evenly between low and high risk customers. We want a way to equalize for underlying riskiness. So we calculate a propensity score to see the likelihood of a customer being enrolled in the PPP. We then look at what ACTUALLY happened (if they were enrolled, and if they charged off). We divide that by the propensity score to ensure that low population outcomes / segments are properly represented/weighted.

1. 𝑃 (𝐴 = 𝑎 ∣ L) = 𝑓(L) = 𝑓(L; 𝛽): Model the probability a customer is enrolled in the PPP, given their covariates. Basically, use a logistic regression where response variable is A (PPP enrollment), and all variables except A and Y are predictors. Results in function that assigns every customer a probability of being enrolled in PPP - propensity score.

2. I(A=a): indicator flagging to only look at people who actually received the treatment we're interested in. So, if A=1, only look at people who were actually enrolled.

3. Y: actual missed payment status, for everyone the indicator flags.
Overall: Someone had a 10% chance of enrolling but actually did enroll (A=1), and did NOT miss their pmt (Y=0): denominator is 0.1, while numerator is 1. So now their missed payment status is 10x the average person in terms of weight.

```{r IPW all var calc ps}
#get propensity scores
formula_ps_all_var <- as.formula(paste("A~", paste(cov_all, collapse = "+"))) #formula of A~all covariates
model_ps_all_var <- glm(formula_ps_all_var, data = data, family = binomial) #binomial because only 2 outcomes

print("formula")
print(formula_ps_all_var)

data$ps_all_var <- predict(model_ps_all_var, type = "response") #calculate propensity scores for all customers; response transforms to probabilities.
print("min and max ps")
print(min(data$ps_all_var)) #.199515e-08
print(max(data$ps_all_var)) #1
#extreme ps means likely have extreme scores

data$ps_all_var_clamped <- pmax(pmin(data$ps_all_var, 0.9999), 0.0001) #clamped ps; if 0 or 1 then model has divide by 0 issues potentially - more apparent in bootstrapping. this is a sign of too many variables.


# calculate weights.
data$weights_all_var <- ifelse(data[["A"]] == 1, 1 / data$ps_all_var_clamped, 1 / (1 - data$ps_all_var_clamped)) 
print("min and max weights")
print(min(data$weights_all_var))
print(max(data$weights_all_var))
#if customer was actually enrolled (A==1), then d weight as 1/ps, otherwise do 1/(1-ps)
#So if customer had A=1 & ps=0.8, then weight = 1.25; but if A=0 then weight=5. Scales by unexpectedness of actual event.

# Estimate ACE using the weighted average
#filter to only people who were enrolled A=1 or A=0
a0_group <- data[data$A == 0, ]
a1_group <- data[data$A == 1, ]
# Filter to A0 or A1 groups, get default and multiply by weight; basically, if paid on time, contributes nothing to numerator. numerator = total # of defaults, denominator = total # of people (or sum of weights here)
EY_a0_ipw_all_var <- sum(as.numeric(as.character(a0_group$Y)) * a0_group$weights_all_var)/sum(a0_group$weights_all_var)
EY_a1_ipw_all_var <- sum(as.numeric(as.character(a1_group$Y)) * a1_group$weights_all_var)/sum(a1_group$weights_all_var)

ace_ipw_all_var <- EY_a1_ipw_all_var - EY_a0_ipw_all_var
print(paste("IPW ACE (Full Set):", round(ace_ipw_all_var,4)))
```



Soem graphs are below just to look at distribution of scores, etc. Given those graphs and the model error from bootstrapping, I had decided to clamp the propensity scores because otherwise we would potentially get a divide by 0 error.

```{r}
# Create a histogram of Propensity Scores
hist(data$ps_all_var, 
     breaks = 50, 
     main = "Distribution of Propensity Scores",
     xlab = "Propensity Score", 
     col = "skyblue", 
     border = "white")
#a lot of low propensity scores
```

```{r}

#Plot by Treatment Group to see "Overlap"
ggplot(data, aes(x = ps_all_var, fill = as.factor(A))) +
  geom_density(alpha = 0.5) +
  labs(title = "Overlap of Propensity Scores",
       x = "Propensity Score",
       fill = "Enrolled (A)") +
  theme_minimal()

```

```{r}
#plot weights
hist(data$weights_all_var, 
     breaks = 50, 
     main = "Distribution of weights",
     xlab = "weights", 
     col = "skyblue", 
     border = "white")
```

```{r}
ggplot(data, aes(x = weights_all_var, fill = as.factor(A))) +
  geom_density(alpha = 0.5) +
  labs(title = "Overlap of Weights",
       x = "Weights",
       fill = "Enrolled (A)") +
  theme_minimal()
#no real takeaway. A=1 has higher weights in general maybe?
```

```{r}
#see how weights are distr A=1 vs A=0
data %>%
  group_by(A) %>%
  summarise(
    Count = n(),
    Min     = min(weights_all_var, na.rm = TRUE),
    Average = mean(weights_all_var, na.rm = TRUE),
    Median  = median(weights_all_var, na.rm = TRUE),
    Perc_25 = quantile(weights_all_var, 0.25, na.rm = TRUE),
    Perc_75 = quantile(weights_all_var, 0.75, na.rm = TRUE),
    Max     = max(weights_all_var, na.rm = TRUE),
    na_count = sum(is.na(weights_all_var)),
    inf_count = sum(is.infinite(weights_all_var))
  )
#although max weight for A=0 is bigger, overall A=1 tends to have higher weights given perc_75, median, and avg are all higher than A=0; perc_25 is similar, though.
#no NAs or inf
```



# All Variables, bootstrap, 

Bootstrap to get the bootstrap bias (unsure if important) and variance!

Going forward, I will do bootstrapping so I can capture variance, etc.

Note that I clamp PS scores below, and throughout the rest.

```{r bootstrap start all var}
bootstrap_results_all_var <- replicate(bootstrap_sample_size, {
  #create bootstrap sample
  sample_df <- data[sample(nrow(data), replace = TRUE), ]

  #outcome regression
  #run model
  formula_or_sample_df <- as.formula(paste("Y ~ A +", paste(cov_all, collapse = "+"))) 
  model_or_sample_df <- glm(formula_or_sample_df, data = sample_df, family = binomial)

  #create fake datasets
  sample_df_a0 <- sample_df
  sample_df_a0$A <- factor(0, levels = levels(sample_df$A))  
  sample_df_a1 <- sample_df
  sample_df_a1$A <- factor(1, levels = levels(sample_df$A)) 
  
  # Predict default probability for world of everyone enrolled, and no one enrolled
  sample_df_prob_Y_if_A1 <- predict(model_or_sample_df, newdata = sample_df_a1, type = "response") 
  EY_a1_or_sample_df <- mean(sample_df_prob_Y_if_A1)

  sample_df_prob_Y_if_A0 <- predict(model_or_sample_df, newdata = sample_df_a0, type = "response") 
  EY_a0_or_sample_df <- mean(sample_df_prob_Y_if_A0)
  
  ace_or_sample_df <- EY_a1_or_sample_df - EY_a0_or_sample_df
  
  #IPW
  #get propensity score
  formula_ps_sample_df <- as.formula(paste("A~", paste(cov_all, collapse = "+")))
  model_ps_sample_df <- glm(formula_ps_sample_df, data = sample_df, family = binomial)
  sample_df$ps_sample_df <- predict(model_ps_sample_df, type = "response")
  #clamp ps score between 0.0001 and 0.9999 because otherwise get warnings of fitted prob being 0 or 1 occurring
  sample_df$ps_sample_df_clamped <- pmax(pmin(sample_df$ps_sample_df, 0.9999), 0.0001)
  #get weights
  sample_df$weights_sample_df <- ifelse(sample_df$A == 1, 1/sample_df$ps_sample_df_clamped, 1/(1-sample_df$ps_sample_df_clamped))
  #divide into a0 and a1 groups
  a0_group_sample_df <- sample_df[sample_df$A == 0, ]
  a1_group_sample_df <- sample_df[sample_df$A == 1, ]
  #calculate ACE
  EY_a0_ipw_sample_df <- sum(as.numeric(as.character(a0_group_sample_df$Y)) * a0_group_sample_df$weights_sample_df)/sum(a0_group_sample_df$weights_sample_df)
  EY_a1_ipw_sample_df <- sum(as.numeric(as.character(a1_group_sample_df$Y)) * a1_group_sample_df$weights_sample_df)/sum(a1_group_sample_df$weights_sample_df)
  ace_ipw_sample_df <-  EY_a1_ipw_sample_df - EY_a0_ipw_sample_df
  return(c(ace_or_sample_df, ace_ipw_sample_df))
})
```

```{r}
# print(bootstrap_results_all_var)
```

Basically, this outputs 2 rows of data ([1,], [2,]). With the 1st row being OR and 2nd being IPW method. and then 10 columns (or however many runs of columns).

```{r bootstrap variance and bias for all variables}
orig_ace_all_var <- c(ace_or_all_var, ace_ipw_all_var)

# convert to dataframe
boot_df_all_var <- as.data.frame(t(bootstrap_results_all_var))
colnames(boot_df_all_var) <- c("Regression", "IPW")

# Calculate Metrics
metrics_all_var <- data.frame(
  Method = c("Regression", "IPW"),
  Original_ACE = orig_ace_all_var, # Use values from the full dataset
  Bootstrap_ACE = c(mean(boot_df_all_var$Regression), mean(boot_df_all_var$IPW)),
  Standard_Error = c(sd(boot_df_all_var$Regression), sd(boot_df_all_var$IPW))
)

# Calculate Bias
metrics_all_var$Bootstrap_Bias <- metrics_all_var$Bootstrap_ACE - metrics_all_var$Original_ACE
print(metrics_all_var)
```

This suggests that regression results in a much more stable result, although it is very unintuitive. IPW results in a much higher variance; even with this variance, it is a more intuitive answer as it still suggests a payment plan decreases default risk. Regression is probably biased by overfitting noise and thus creating bias (no proof here, just judgemental unintuitiveness), while IPW could be distorted by extreme weights (as evidenced by high standard error).





# 2 Now the Analysis, for lasso-selected variables!




# Lasso Variables, Outcome regression , not transformed

```{r outcome reg lasso vars}
#set up the data
x_or_lasso <- model.matrix(Y ~ A + ., data = data[, c("Y", "A", cov_all)])[,-1] #only pull in cov_all, Y, A. Transform factorized variables into dummy-encoded variables. Uses everything pulled in as a predictor. [-1] Removes intercept automatically included by model.matrix
y_or_lasso <- data$Y
print(colnames(x_or_lasso))

col_idx_A1 <- which(colnames(x_or_lasso) == "A1")
p_fac_or <- rep(1, ncol(x_or_lasso))
p_fac_or[col_idx_A1] <- 0

#for A in (penalty factor 0)
#fit lasso with cross-validation, get best lambda
model_or_lasso <- cv.glmnet(x_or_lasso, y_or_lasso, family = "binomial", alpha = 1, penalty.factor = p_fac_or)
best_lambda_or_lasso <- model_or_lasso$lambda.min
print(coef(model_or_lasso))

# Predict default probability for world of everyone enrolled, and no one enrolled

#basically, A1 is a dummy column for the factor A=1
#get the column number for that
# create new datasets where A is overridden to be all 1s or all 0s, and then transform into model matrix format for prediction
x_all_enrolled_or_lasso <- x_or_lasso
x_all_enrolled_or_lasso[, col_idx_A1] <- 1  # Everyone treated
x_no_enrolled_or_lasso <- x_or_lasso
x_no_enrolled_or_lasso[, col_idx_A1] <- 0  # Everyone untreated

#now predict
data$prob_Y_if_A1_lasso <- as.numeric(predict(model_or_lasso, s = best_lambda_or_lasso, newx = x_all_enrolled_or_lasso, type = "response"))
EY_a1_or_lasso <- mean(data$prob_Y_if_A1_lasso)

data$prob_Y_if_A0_lasso <- as.numeric(predict(model_or_lasso, s = best_lambda_or_lasso, newx = x_no_enrolled_or_lasso, type = "response"))
EY_a0_or_lasso <- mean(data$prob_Y_if_A0_lasso)


ace_or_lasso <- EY_a1_or_lasso - EY_a0_or_lasso

print("formula")
lasso_y_coefs <- coef(model_or_lasso, s = "lambda.min")
lasso_y_vars <- rownames(lasso_y_coefs)[which(as.matrix(lasso_y_coefs) != 0)]
# Remove Intercept and Treatment A to get the adjustment set
cov_lasso_y <- setdiff(lasso_y_vars, c("(Intercept)", "A"))
formula_lasso_or_text <- paste("Y ~ ", paste(cov_lasso_y, collapse = " + "))
formula_lasso_or_text_anotherver <- lasso_y_vars[which(lasso_y_vars != "(Intercept)")]
print(formula_lasso_or_text)
print(lasso_y_vars)

#print means
print("Default likelihood (Lasso, OR), A=1 then A=0")
print(EY_a1_or_lasso)
print(EY_a0_or_lasso)
print(paste("OR ACE (Lasso):", round(ace_or_lasso,4)))
```




# Lasso Variables, IPW


```{r IPW lasso var calc ps}


#prepare data
x_ps_lasso <- model.matrix(A ~ ., data = data[, c("A", cov_all)])[,-1]
y_ps_lasso <- data$A


#fit lasso
model_ps_lasso <- cv.glmnet(x_ps_lasso, y_ps_lasso, family = "binomial", alpha = 1)
best_lambda_ps_lasso <- model_ps_lasso$lambda.min


#get propensity scores
data$ps_lasso <- as.numeric(predict(model_ps_lasso, s = best_lambda_ps_lasso, newx = x_ps_lasso, type = "response"))


print("min and max ps")
print(min(data$ps_lasso))
print(max(data$ps_lasso))


data$ps_lasso_clamped <- pmax(pmin(data$ps_lasso, 0.9999), 0.0001)


# calculate weights.
data$weights_lasso <- ifelse(data[["A"]] == 1, 1 / data$ps_lasso_clamped, 1 / (1 - data$ps_lasso_clamped))


print("min and max weights")
print(min(data$weights_lasso))
print(max(data$weights_lasso))


#get formula
print("formula")
lasso_a_coefs <- coef(model_ps_lasso, s = "lambda.min")
lasso_a_vars <- rownames(lasso_a_coefs)[which(as.matrix(lasso_a_coefs) != 0)]
cov_lasso_a <- setdiff(lasso_a_vars, c("(Intercept)"))
formula_lasso_ps_text <- paste("A ~", paste(cov_lasso_a, collapse = " + "))
formula_lasso_ps_text_anotherver <- lasso_a_vars[which(lasso_a_vars != "(Intercept)")]
print(formula_lasso_ps_text)
print(cov_lasso_a)

# Estimate ACE using the weighted average
#filter to only people who were enrolled A=1 or A=0
a0_group <- data[data$A == 0, ]
a1_group <- data[data$A == 1, ]
# Filter to A0 or A1 groups, get default and multiply by weight
EY_a0_ipw_lasso <- sum(as.numeric(as.character(a0_group$Y))* a0_group$weights_lasso)/sum(a0_group$weights_lasso)
EY_a1_ipw_lasso <- sum(as.numeric(as.character(a1_group$Y)) * a1_group$weights_lasso)/sum(a1_group$weights_lasso)


ace_ipw_lasso <- EY_a1_ipw_lasso - EY_a0_ipw_lasso
print(paste("IPW ACE (Lasso):", round(ace_ipw_lasso,4)))
```


```{r}
#see how weights are distr A=1 vs A=0
data %>%
 group_by(A) %>%
 summarise(
   Count = n(),
   Min     = min(weights_lasso, na.rm = TRUE),
   Average = mean(weights_lasso, na.rm = TRUE),
   Median  = median(weights_lasso, na.rm = TRUE),
   Perc_25 = quantile(weights_lasso, 0.25, na.rm = TRUE),
   Perc_75 = quantile(weights_lasso, 0.75, na.rm = TRUE),
   Max     = max(weights_lasso, na.rm = TRUE),
   na_count = sum(is.na(weights_lasso)),
   inf_count = sum(is.infinite(weights_lasso))
 )
#although max weight for A=0 is bigger, overall A=1 tends to have higher weights given perc_75, median, and avg are all higher than A=0; perc_25 is similar, though.
#no NAs or inf
```


This shows the weights are more reasonable than the full dataset.



# Lasso Variables, bootstrap


```{r bootstrap start lasso}
bootstrap_results_lasso <- replicate(bootstrap_sample_size, {
 #create bootstrap sample
 sample_df <- data[sample(nrow(data), replace = TRUE), ]

 #outcome regression

 x_or_lasso_sample_df <- model.matrix(Y ~ A + ., data = sample_df[, c("Y", "A", cov_all)])[,-1]
 y_or_lasso_sample_df <- sample_df$Y

 col_idx_A1_sample_df <- which(colnames(x_or_lasso_sample_df) == "A1")
 p_fac_or_sample_df <- rep(1, ncol(x_or_lasso_sample_df))
 p_fac_or_sample_df[col_idx_A1_sample_df] <- 0

 #fit lasso with cross-validation, get best lambda
 model_or_lasso_sample_df <- cv.glmnet(x_or_lasso_sample_df, y_or_lasso_sample_df, family = "binomial", alpha = 1, penalty.factor = p_fac_or_sample_df)
 best_lambda_or_lasso_sample_df <- model_or_lasso_sample_df$lambda.min


 x_all_enrolled_or_lasso_sample_df <- x_or_lasso_sample_df
 x_all_enrolled_or_lasso_sample_df[, col_idx_A1_sample_df] <- 1  # Everyone treated
 x_no_enrolled_or_lasso_sample_df <- x_or_lasso_sample_df
 x_no_enrolled_or_lasso_sample_df[, col_idx_A1_sample_df] <- 0  # Everyone untreated

 #now predict
 sample_df_prob_Y_if_A1 <- as.numeric(predict(model_or_lasso_sample_df, s = best_lambda_or_lasso_sample_df, newx = x_all_enrolled_or_lasso_sample_df, type = "response"))
 EY_a1_or_lasso_sample_df <- mean(sample_df_prob_Y_if_A1)


 sample_df_prob_Y_if_a0 <- as.numeric(predict(model_or_lasso_sample_df, s = best_lambda_or_lasso_sample_df, newx = x_no_enrolled_or_lasso_sample_df, type = "response"))
 EY_a0_or_lasso_sample_df <- mean(sample_df_prob_Y_if_a0)


 ace_or_lasso_sample_df <- EY_a1_or_lasso_sample_df - EY_a0_or_lasso_sample_df


  #IPW


 #prepare data
 x_ps_lasso_sample_df <- model.matrix(A ~ ., data = sample_df[, c("A", cov_all)])[,-1]
 y_ps_lasso_sample_df <- sample_df$A


 #fit lasso
 model_ps_lasso_sample_df <- cv.glmnet(x_ps_lasso_sample_df, y_ps_lasso_sample_df, family = "binomial", alpha = 1)
 best_lambda_ps_lasso_sample_df <- model_ps_lasso_sample_df$lambda.min
 #get propensity scores
 sample_df$ps_lasso_sample_df <- as.numeric(predict(model_ps_lasso_sample_df, s = best_lambda_ps_lasso_sample_df, newx = x_ps_lasso_sample_df, type = "response"))


 sample_df$ps_lasso_clamped_sample_df <- pmax(pmin(sample_df$ps_lasso_sample_df, 0.9999), 0.0001)
 # calculate weights.
 sample_df$weights_lasso_sample_df <- ifelse(sample_df[["A"]] == 1, 1 / sample_df$ps_lasso_clamped_sample_df, 1 / (1 - sample_df$ps_lasso_clamped_sample_df))


 # Estimate ACE using the weighted average
 #filter to only people who were enrolled A=1 or A=0
 a0_group_sample_df <- sample_df[sample_df$A == 0, ]
 a1_group_sample_df <- sample_df[sample_df$A == 1, ]
 # Filter to A0 or A1 groups, get default and multiply by weight
 EY_a0_ipw_lasso_sample_df <- sum(as.numeric(as.character(a0_group_sample_df$Y)) * a0_group_sample_df$weights_lasso_sample_df)/sum(a0_group_sample_df$weights_lasso_sample_df)
 EY_a1_ipw_lasso_sample_df <- sum(as.numeric(as.character(a1_group_sample_df$Y)) * a1_group_sample_df$weights_lasso_sample_df)/sum(a1_group_sample_df$weights_lasso_sample_df)


 ace_ipw_lasso_sample_df <-  EY_a1_ipw_lasso_sample_df - EY_a0_ipw_lasso_sample_df


 return(c(ace_or_lasso_sample_df, ace_ipw_lasso_sample_df))
})
```


```{r bootstrap variance and bias for lasso}
orig_ace_lasso <- c(ace_or_lasso, ace_ipw_lasso)


# convert to dataframe
boot_df_lasso <- as.data.frame(t(bootstrap_results_lasso))
colnames(boot_df_lasso) <- c("Regression", "IPW")


# Calculate Metrics
metrics_lasso <- data.frame(
 Method = c("Regression", "IPW"),
 Original_ACE = orig_ace_lasso, # Use values from the full dataset
 Bootstrap_ACE = c(mean(boot_df_lasso$Regression), mean(boot_df_lasso$IPW)),
 Standard_Error = c(sd(boot_df_lasso$Regression), sd(boot_df_lasso$IPW))
)


# Calculate Bias
metrics_lasso$Bootstrap_Bias <- metrics_lasso$Bootstrap_ACE - metrics_lasso$Original_ACE
print(metrics_lasso)
```


This suggests that things are more stable with the lasso.





# 3 Now, do rank PC analysis!!

PC requires data follow Gaussian distribution and have linear relationships - that's why we transformed the variables. The "Rank PC" method aims to make the PC algorithm robust to non-normal distributions and non-linear relationships.

PC systematically tests for conditional independence between variables to determine if there should be an edge - gaussian conditional independence test, assuming variables normally distributed. 

Rank PC uses rank based test, say using spearman's rank correlation (considers rank of value not raw values). Does not look at raw values - looks at rank of the value relative to others. Does not assume a specific distribution. Somewhat robust to outliers, because just ranking. On the other hand, if data is not evenly spread, could misrepresent? Because it treats gaps of 100, 101, and 102 as the same as 100, 5k, 100k. 

# Non-transformed, Rank PC

```{r obtain DAG rank pc}
#convert data to numerical for gaussian tests
rank_pc_data <- data %>%
  mutate(across(where(is.factor), ~as.numeric(as.factor(.x)))) %>% #rank my factors into numbers
  select(Y, A, all_of(cov_all))

# calculate correlation matrix and sample size; use to test for conditional independence amongst pairs
rank_pc_corr_size <- list(C = cor(rank_pc_data, method = "spearman"), n = nrow(rank_pc_data))

# Run PC; alpha serves as a tuning parameter for the significance of independence tests
rank_pc_fit <- pc(suffStat = rank_pc_corr_size,
                    indepTest = gaussCItest,
                    labels = colnames(rank_pc_data),
                    alpha = 0.01, 
                    verbose = FALSE)

# Plot DAG
plot(rank_pc_fit@graph, main = "Causal DAG: Rank PC Method")

#export
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_rank_pc_method.png", width = 1200, height = 1000, res = 150)

plot(rank_pc_fit@graph, main = "Causal DAG: Rank PC Method")

# 3. Close the device to save the file
dev.off()
```


```{r obtain markov blanket rank pc }

rank_pc_dag <- pcalg2dagitty(as(rank_pc_fit@graph, "matrix"), 
                       labels = colnames(rank_pc_data), 
                       type = "cpdag")

# Get the Markov Blanket (Parents + Children + Spouses)
rank_pc_mb_y <- markovBlanket(rank_pc_dag, "Y")
print(rank_pc_mb_y)

# Update your covariate list for the models
cov_rank_pc_y <- setdiff(rank_pc_mb_y, "A") #exclude A
print(cov_rank_pc_y) #only PAY_0?
```

```{r}
# Get the Markov Blanket (Parents + Children + Spouses)
rank_pc_mb_a <- markovBlanket(rank_pc_dag, "A")
print(rank_pc_mb_a)

# Update your covariate list for the models
cov_rank_pc_a <- setdiff(rank_pc_mb_a, "Y") #exclude A
print(cov_rank_pc_a) #only PAY_0?
```
```


# rank pc algorithm nontransformed or
```{r run or on rank pc}

#outcome regression
if(length(cov_rank_pc_y) == 0) {
  formula_or_rank_pc <- as.formula("Y ~ A")
} else {
  formula_or_rank_pc <- as.formula(paste("Y ~ A +", paste(cov_rank_pc_y, collapse = "+")))
}
model_or_rank_pc <- glm(formula_or_rank_pc, data = data, family = binomial)

print("formula")
print(formula_or_rank_pc)

# pc_data_all_enrolled <- data_all_enrolled %>%
#   # mutate(across(everything(), as.numeric)) %>%
#   select(Y, A, all_of(cov_all))

# pc_data_no_enrolled <- data_no_enrolled %>%
#   # mutate(across(everything(), as.numeric)) %>%
#   select(Y, A, all_of(cov_all))

# Predict default probability for world of everyone enrolled, and no one enrolled
data$prob_Y_if_A1_rank_pc_var<- predict(model_or_rank_pc, newdata = data_all_enrolled, type = "response") #type=response gives 0-1 prob

EY_a1_or_rank_pc_var <- mean(data$prob_Y_if_A1_rank_pc_var)

data$prob_Y_if_A0_rank_pc_var <- predict(model_or_rank_pc, newdata = data_no_enrolled, type = "response")
EY_a0_or_rank_pc_var <- mean(data$prob_Y_if_A0_rank_pc_var)

ace_or_rank_pc_var <- EY_a1_or_rank_pc_var - EY_a0_or_rank_pc_var

#print means
print("Default likelihood (Rank PC Set, OR), A=1 then A=0")
print(EY_a1_or_rank_pc_var)
print(EY_a0_or_rank_pc_var)

print(paste("OR ACE (Rank PC Set):", round(ace_or_rank_pc_var,4))) 
```


# Rank PC algorithm nonttransformed ipw

```{r IPW rank pc var calc ps }

if(length(cov_rank_pc_a) == 0) {
  formula_ps_rank_pc_var <- as.formula("A ~ 1")
} else {
  formula_ps_rank_pc_var <- as.formula(paste("A~", paste(cov_rank_pc_a, collapse = "+"))) #formula of A~pc covariates
}
#get propensity scores
model_ps_rank_pc_var <- glm(formula_ps_rank_pc_var, data = data, family = binomial) #binomial because only 2 outcomes

print("formula")
print(formula_ps_rank_pc_var)

data$ps_rank_pc_var <- predict(model_ps_rank_pc_var, type = "response") #calculate propensity scores for all customers; response transforms to probabilities.
print("min and max ps")
print(min(data$ps_rank_pc_var)) 
print(max(data$ps_rank_pc_var)) 

#extreme ps means likely have extreme scores

data$ps_rank_pc_var_clamped <- pmax(pmin(data$ps_rank_pc_var, 0.9999), 0.0001) 


# calculate weights.
data$weights_rank_pc_var <- ifelse(data[["A"]] == 1, 1 / data$ps_rank_pc_var_clamped, 1 / (1 - data$ps_rank_pc_var_clamped)) 

print("min and max weights")
print(min(data$weights_rank_pc_var)) #1.0001
print(max(data$weights_rank_pc_var)) #2.8


#if customer was actually enrolled (A==1), then d weight as 1/ps, otherwise do 1/(1-ps)
#So if customer had A=1 & ps=0.8, then weight = 1.25; but if A=0 then weight=5. Scales by unexpectedness of actual event.

# Estimate ACE using the weighted average
#filter to only people who were enrolled A=1 or A=0
a0_group <- data[data$A == 0, ]
a1_group <- data[data$A == 1, ]
# Filter to A0 or A1 groups, get default and multiply by weight; basically, if paid on time, contributes nothing to numerator. numerator = total # of defaults, denominator = total # of people (or sum of weights here)
EY_a0_ipw_rank_pc_var <- sum(as.numeric(as.character(a0_group$Y)) * a0_group$weights_rank_pc_var)/sum(a0_group$weights_rank_pc_var)
EY_a1_ipw_rank_pc_var <- sum(as.numeric(as.character(a1_group$Y)) * a1_group$weights_rank_pc_var)/sum(a1_group$weights_rank_pc_var)

ace_ipw_rank_pc_var <- EY_a1_ipw_rank_pc_var - EY_a0_ipw_rank_pc_var
print(paste("IPW ACE (Rank PC set):", round(ace_ipw_rank_pc_var,4)))
```


# Rank Pc algorithm bootstrap nontransformed
```{r bootstrap rank pc method nontransformed}
bootstrap_results_rank_pc <- replicate(bootstrap_sample_size, {
  #bootstrap sample
  sample_df <- data[sample(nrow(data), replace = TRUE), ]
  
  #create pc markov blanket
  boot_data <- sample_df %>% 
  mutate(across(where(is.factor), ~as.numeric(as.factor(.x)))) %>% #rank my factors into numbers
  select(Y, A, all_of(cov_all))
  
  #get correlation, sample size of bootstrapped data
  suff_boot <- list(C = cor(boot_data, method = "spearman"), n = nrow(boot_data))
  
  #Run rank PC Algorithm
  fit_boot <- pc(suffStat = suff_boot, indepTest = gaussCItest,
                    labels = colnames(boot_data), alpha = 0.01, verbose = FALSE)

  adj_matrix <- as(fit_boot@graph, "matrix")
  adj_matrix_input <- pcalg2dagitty(adj_matrix, 
                        labels = colnames(boot_data), 
                        type = "cpdag")

  # Get the Markov Blanket (Parents + Children + Spouses)
  y_indices_all_boot <- markovBlanket(adj_matrix_input, "Y")
  cov_boot <- setdiff(y_indices_all_boot, "A") #exclude A

  a_indices_all_boot <- markovBlanket(adj_matrix_input, "A")
  cov_boot_a <- setdiff(a_indices_all_boot, "Y") 

  #what is markov blanket empty?
  if(length(cov_boot) == 0) {
    formula_or_boot <- as.formula("Y ~ A")
  } else {
    formula_or_boot <- as.formula(paste("Y ~ A +", paste(cov_boot, collapse = "+")))
  }
  if(length(cov_boot_a) == 0) {
    formula_ps_boot <- as.formula("A ~ 1")
  } else {
    formula_ps_boot <- as.formula(paste("A ~", paste(cov_boot_a, collapse = "+")))
  }
  
  # outcome regression
  model_or_boot <- glm(formula_or_boot, data = sample_df, family = binomial)
  
  #create fake datasets
  sample_df_a0 <- sample_df
  sample_df_a0$A <- factor(0, levels = levels(sample_df$A))
  sample_df_a1 <- sample_df
  sample_df_a1$A <- factor(1, levels = levels(sample_df$A))

  # Predict default probability for world of everyone enrolled, and no one enrolled
  sample_df_prob_Y_if_A1 <- predict(model_or_boot, newdata = sample_df_a1, type = "response") 
  EY_a1_or_sample_df <- mean(sample_df_prob_Y_if_A1)

  sample_df_prob_Y_if_A0 <- predict(model_or_boot, newdata = sample_df_a0, type = "response") 
  EY_a0_or_sample_df <- mean(sample_df_prob_Y_if_A0)
  
  ace_or_sample_df <- EY_a1_or_sample_df - EY_a0_or_sample_df
  
  #IPW
  #get propensity score
  model_ps_sample_df <- glm(formula_ps_boot, data = sample_df, family = binomial)
  sample_df$ps_sample_df <- predict(model_ps_sample_df, type = "response")
  #clamp ps score between 0.0001 and 0.9999 because otherwise get warnings of fitted prob being 0 or 1 occurring
  sample_df$ps_sample_df_clamped <- pmax(pmin(sample_df$ps_sample_df, 0.9999), 0.0001)
  #get weights
  sample_df$weights_sample_df <- ifelse(sample_df$A == 1, 1/sample_df$ps_sample_df_clamped, 1/(1-sample_df$ps_sample_df_clamped))
  #divide into a0 and a1 groups
  a0_group_sample_df <- sample_df[sample_df$A == 0, ]
  a1_group_sample_df <- sample_df[sample_df$A == 1, ]
  #calculate ACE
  EY_a0_ipw_sample_df <- sum(as.numeric(as.character(a0_group_sample_df$Y)) * a0_group_sample_df$weights_sample_df)/sum(a0_group_sample_df$weights_sample_df)
  EY_a1_ipw_sample_df <- sum(as.numeric(as.character(a1_group_sample_df$Y)) * a1_group_sample_df$weights_sample_df)/sum(a1_group_sample_df$weights_sample_df)
  ace_ipw_sample_df <-  EY_a1_ipw_sample_df - EY_a0_ipw_sample_df

  return(list(ace_or = ace_or_sample_df, 
              ace_ipw = ace_ipw_sample_df, 
              adj_matrix = adj_matrix,
              blanket = y_indices_all_boot,
              blanket_a = a_indices_all_boot
              ))
}, simplify = FALSE)
```



```{r bootstrap variance and bias for rank pc variables}
orig_ace_rank_pc_var <- c(ace_or_rank_pc_var, ace_ipw_rank_pc_var)

#extract ACE values
boot_aces_or_rank_pc  <- unlist(lapply(bootstrap_results_rank_pc, function(x) x$ace_or))
boot_aces_ipw_rank_pc <- unlist(lapply(bootstrap_results_rank_pc, function(x) x$ace_ipw))

#extract markov blanket
boot_blankets_rank_pc <- lapply(bootstrap_results_rank_pc, function(x) x$blanket)
boot_blankets_rank_pc_a <- lapply(bootstrap_results_rank_pc, function(x) x$blanket_a)

# convert to dataframe
boot_df_rank_pc_var <- data.frame(
  Regression = boot_aces_or_rank_pc,
  IPW = boot_aces_ipw_rank_pc
)
colnames(boot_df_rank_pc_var) <- c("Regression", "IPW")

# Calculate Metrics
metrics_rank_pc_var <- data.frame(
  Method = c("Regression", "IPW"),
  Original_ACE = orig_ace_rank_pc_var, # Use values from the full dataset
  Bootstrap_ACE = c(mean(boot_df_rank_pc_var$Regression), mean(boot_df_rank_pc_var$IPW)),
  Standard_Error = c(sd(boot_df_rank_pc_var$Regression), sd(boot_df_rank_pc_var$IPW))
)

# Calculate Bias
metrics_rank_pc_var$Bootstrap_Bias <- metrics_rank_pc_var$Bootstrap_ACE - metrics_rank_pc_var$Original_ACE
print(metrics_rank_pc_var)
```


```{r markov blankets rank pc}
boot_blankets_rank_pc_unlisted <- unlist(boot_blankets_rank_pc)

boot_mb_stability_rank_pc <- table(boot_blankets_rank_pc_unlisted) / bootstrap_sample_size
print("Selection Frequency for Y's Markov Blanket:")
print(sort(boot_mb_stability_rank_pc, decreasing = TRUE))
```

```{r markov blankets rank pc a}
boot_blankets_rank_pc_unlisted_a <- unlist(boot_blankets_rank_pc_a)

boot_mb_stability_rank_pc_a <- table(boot_blankets_rank_pc_unlisted_a) / bootstrap_sample_size
print("Selection Frequency for A's Markov Blanket:")
print(sort(boot_mb_stability_rank_pc_a, decreasing = TRUE))
```

```{r DAG avg rank pc, fig.width=10, fig.height=8, out.width="100%"}
#get adjacent matrices
all_adj_bootstrap_rank_pc <- lapply(bootstrap_results_rank_pc, function(x) x$adj_matrix)
#sum them up to get edge frequencies - prob an edge exists between 2 variables
consensus_matrix_bootstrap_rank_pc <- Reduce("+", all_adj_bootstrap_rank_pc) / bootstrap_sample_size #reduce sums all matrices. if arrow from pay_0 to y is in every bootstrap, then it will have bootstrap_sample_size values in that matrix cell. if only half the time, then half the size.

# Create a graph where an edge exists if it appeared in > 50% of samples
average_adj_bootstrap_rank_pc <- (consensus_matrix_bootstrap_rank_pc > 0.5) * 1

# Convert back to a graph object for plotting
avg_graph_bootstrap_rank_pc <- graph_from_adjacency_matrix(average_adj_bootstrap_rank_pc, mode = "directed")
plot(avg_graph_bootstrap_rank_pc, main = "Consensus DAG (Edges in >50% of Bootstraps)")

png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/consensus_dag_rank_pc_bootstrap.png", width = 1200, height = 1000, res = 150)
par(mar = c(1, 1, 3, 1)) 
plot(avg_graph_bootstrap_rank_pc, 
     main = "Consensus DAG (Edges in >50% of Bootstraps): Rank PC Method",
     vertex.size = 15,              # Smaller nodes
     vertex.label.cex = 0.6,        # Smaller labels
     vertex.label.dist = 1.2,       # Offset labels from node centers
     edge.arrow.size = 0.3,         # Smaller arrows
     edge.curved = 0.2)             # Slight curve to distinguish overlapping arrows
dev.off()
```






# 4 Now the Analysis, for PC-selected variables! (limited vars to most recent month first)


# PC Algorithm

Find the markov blanket of Y, consisting of its parents, children, and parents of its children. Y is conditionally independent of all other variables GIVEN its Markov Blanket. 

So: use PC algorithm to find the full DAG. Identify variables directly connected to Y. Run OR and IPW using the variables directly connected to Y.

Note: assumes
1. Causal sufficiency - no hidden confounders
2. faithfulness: conditional independencies in the data imply cond ind in the DAG.
3. Markov condition: variable is conditionally indpendent of its non-descendents given its parents, in the DAG.

PC measures cond ind in the data, and searches for a graph to satisfy them.

1. start with fully connected, UNDIRECTED graph
2. test for conditional independence amongst all pairs nad subsets of variables --> X ind of Y given Z, for different sets of Z
3. remove edges based on these tests
4. continue until we can't remove more edges without violating conditional independences in data
5. apply orientation rules (colliders etc) to orient undirected edges, as possible
6. adjust to make sure graph remains acyclic

For Gaussian data, this is like testing if the null hypothesis is rho of X&Y given Z is 0, vs not.
If data is categorical, can test conditional independence using a chi-square test.

Obtain the DAG.

The "Penalty" of Categorical Data

When you use a Gaussian test (like gaussCItest), the algorithm treats everything as a simple linear correlation. It only needs to find one "slope" to connect two variables.

mixCItest is different. When it tests a factor (like EDUCATION with 4 levels) against your outcome Y:
    It essentially creates a contingency table or a grouped model.
    If you only have 500 rows, the data gets split into very small "buckets" (e.g., how many people with "Graduate School" education actually "Defaulted"?).
    If those buckets are too small, the p-value becomes very large (insignificant), and the algorithm deletes the line.
2. Loss of Power in Conditional Tests
The PC algorithm doesn't just test A vs B. It tests A vs B given C, D, and E.
    As the "Conditioning Set" grows, mixCItest has to slice your 500 rows even thinner.
    It is much harder to prove that LIMIT_BAL and Y are dependent after accounting for PAY_0 and EDUCATION when using a mixed-type test. The test "gives up" more easily than a purely numeric one would.
3. "Conservative" by Design
The micd package (which provides mixCItest) is designed to avoid False Positives. It would rather tell you two variables are independent than show a "fake" causal link. In biostatistics and high-stakes financial modeling, a "Sparse" graph is often considered more trustworthy than a "Spaghetti" graph where everything is connected to everything.

```{r obtain DAG pc}
pc_data <- data %>%
  mutate(across(where(is.integer), as.numeric)) %>% #convert all integers to numeric, to run
  select(Y, A, all_of(cov_all))
# pc_data_var_types <- ifelse(sapply(pc_data, is.factor), 1, 0)
print(summary(pc_data)) #proves data is in right format.
# attr(pc_data, "adaptDF") <- TRUE

# Run PC; alpha serves as a tuning parameter for the significance of independence tests
pc_fit_test <- pc(suffStat = pc_data,
                    indepTest = mixCItest,
                    labels = colnames(pc_data),
                    alpha = 0.01, 
                    verbose = FALSE,
                    u2pd = "relaxed", 
                    skel.method = "stable" ,  #order of test doesnt matter
                    maj.rule = FALSE, #if some say a-->b and some say b-->a, then pick direction with more evidence; if tie, then undirected edge. Setting to false means that in a tie, we just pick a direction (not undirected edge)
                    solve.confl = FALSE #if some say a-->b and some say b-->a, then pick a direction
                    )

# Plot DAG
plot(pc_fit_test@graph, main = "TEST Causal DAG: Markov Blanket Discovery (PC Method)")

#export
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_test.png", width = 1200, height = 1000, res = 150)

plot(pc_fit_test@graph, main = "TEST Causal DAG: Markov Blanket Discovery (PC Method)")

# 3. Close the device to save the file
dev.off()
```


```{r obtain DAG pc v2}
# Run PC; alpha serves as a tuning parameter for the significance of independence tests
pc_fit_test <- pc(suffStat = pc_data,
                    indepTest = mixCItest,
                    labels = colnames(pc_data),
                    alpha = 0.01, 
                    verbose = FALSE,
                    u2pd = "relaxed", 
                    skel.method = "stable" ,  #order of test doesnt matter
                    maj.rule = FALSE,
                    solve.confl = TRUE #if some say a-->b and some say b-->a, then pick a direction
                    )

# Plot DAG
plot(pc_fit_test@graph, main = "TEST Causal DAG: Markov Blanket Discovery (PC Method)")

#export
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_v2_test.png", width = 1200, height = 1000, res = 150)

plot(pc_fit_test@graph, main = "TEST Causal DAG: Markov Blanket Discovery (PC Method)")

# 3. Close the device to save the file
dev.off()
```


```{r}
corrplot(cor(model.matrix(~.-1, data=pc_data)), method="circle")

png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/correlation_matrix_default_test.png", width = 1000, height = 1000, res = 150)
corrplot(cor(model.matrix(~.-1, data=pc_data)), method="circle")
dev.off()
```

Note the correlation is really low, resulting in the PC algorithm not finding many edges. This is likely because of the categorical variables- the mixCItest is struggling to find significant relationships because of the small effective sample size when conditioning on multiple variables. This is not solved even when we limit to max conditioning set size of 2.

```{r obtain DAG pc trimmed}
#convert data to numerical for gaussian tests
pc_data <- data %>%
  mutate(across(where(is.integer), as.numeric)) %>% #convert all integers to numeric, to run
  select(Y, A, all_of(cov_all))

# calculate correlation matrix and sample size; use to test for conditional independence amongst pairs
# pc_corr_size <- list(C = cor(pc_data), n = nrow(pc_data))


pc_fit_bnlearn <- pc.stable(pc_data, 
                    # blacklist = master_blacklist, 
                      # test = "mix", #not needed, will automatically detect which test based on variable type
                    alpha = 0.01)


# Get adjacency matrix from PC fit
# pc_fit_adj_matrix <- amat(pc_fit)
# print(pc_fit_adj_matrix)

pc_fit_bnlearn_cextend <- cextend(pc_fit_bnlearn) #directs edges as much as possible, but doesn't orient all edges. Still has some undirected edges.


pc_fit_bnlearn_maxsx <- pc.stable(pc_data, 
                    # blacklist = master_blacklist, 
                      # test = "mix", #not needed, will automatically detect which test based on variable type
                    alpha = 0.01,
                    max.sx=2) #limits size of conditioning set to 2, which should help with the power issue of mixCItest, but also means we won't find as many independencies and thus will have a more connected graph. 
pc_fit_bnlearn_maxsx_cextend <- cextend(pc_fit_bnlearn_maxsx) #directs edges as much as possible, but doesn't orient all edges. Still has some undirected edges.

cov_most_recent <- c("PAY_0", "EDUCATION", "AGE", "SEX", "PAY_AMT1", "BILL_AMT1", "LIMIT_BAL")

pc_data_limitedcov <- pc_data %>%
  select(Y, A, all_of(cov_most_recent))
pc_fit_bnlearn_limitedcov <- pc.stable(pc_data_limitedcov, 
                    # blacklist = master_blacklist, 
                      # test = "mix", #not needed, will automatically detect which test based on variable type
                    alpha = 0.01)
```

```{r}

# Plot DAG
graphviz.plot(pc_fit_bnlearn, shape = "ellipse",
     main = "TEST Full Causal DAG, PC Method")

# 2. Export a high-quality version
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_recentmonth_test.png", width = 1600, height = 1200, res = 150)
graphviz.plot(pc_fit_bnlearn, shape = "ellipse", 
     main = "TEST Full Causal DAG, PC Method")
# 3. Close the device to save the file
dev.off()
```

```{r}
# # Plot DAG
# graphviz.plot(pc_fit_cextend, shape = "ellipse",
#      main = "Full Causal DAG, PC Method")

# # 2. Export a high-quality version
# png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_bnlearn_cextend.png", width = 1600, height = 1200, res = 150)
# graphviz.plot(pc_fit_cextend, shape = "ellipse", 
#      main = "Full Causal DAG, PC Method")
# # 3. Close the device to save the file
# dev.off()

```

```{r}

# Plot DAG
graphviz.plot(pc_fit_bnlearn_maxsx, shape = "ellipse",
     main = "Full Causal DAG, PC Method")

# 2. Export a high-quality version
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_bnlearn_maxsx_test.png", width = 1600, height = 1200, res = 150)
graphviz.plot(pc_fit_bnlearn_maxsx, shape = "ellipse", 
     main = "TEST Full Causal DAG, PC Method")
# 3. Close the device to save the file
dev.off()
```

```{r}
# Plot DAG
graphviz.plot(pc_fit_bnlearn_maxsx_cextend, shape = "ellipse",
     main = "TEST Full Causal DAG, PC Method")

# 2. Export a high-quality version
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_bnlearn_cextend_maxsx_test.png", width = 1600, height = 1200, res = 150)
graphviz.plot(pc_fit_bnlearn_maxsx_cextend, shape = "ellipse", 
     main = "TEST Full Causal DAG, PC Method")
# 3. Close the device to save the file
dev.off()

```


```{r}
# Plot DAG
graphviz.plot(pc_fit_bnlearn_maxsx_cextend, shape = "ellipse",
     main = "TEST Full Causal DAG, PC Method")

# 2. Export a high-quality version
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_bnlearn_cextend_maxsx_test.png", width = 1600, height = 1200, res = 150)
graphviz.plot(pc_fit_bnlearn_maxsx_cextend, shape = "ellipse", 
     main = "TEST Full Causal DAG, PC Method")
# 3. Close the device to save the file
dev.off()

```

Since we see this isn't working well and the DAG is very sparse, let's try just limiting to the most recent variables, which should have stronger relationships with Y and thus be easier for the PC algorithm to find edges.


```{r}

# Plot DAG
graphviz.plot(pc_fit_bnlearn_limitedcov, shape = "ellipse",
     main = "Causal DAG: PC Method, Limited Covariates")

# 2. Export a high-quality version
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_bnlearn_limitedcov.png", width = 1600, height = 1200, res = 150)
graphviz.plot(pc_fit_bnlearn_limitedcov, shape = "ellipse", 
     main = "Causal DAG: PC Method, Limited Covariates")
# 3. Close the device to save the file
dev.off()
```



```{r obtain markov blanket pc }
# Find indices of variables connected to Y
pc_y_indices <- bnlearn::nbr(pc_fit_bnlearn_limitedcov, "Y") #all neighbors - but not spouses
print(pc_y_indices)

#parents, children, spouses using cextend version of the graph, which orients some edges and thus allows us to differentiate between parents and children (but still has some undirected edges, so not perfect)
pc_y_indices_parents <- bnlearn::parents(pc_fit_bnlearn_limitedcov, "Y")
pc_y_indices_children <- bnlearn::children(pc_fit_bnlearn_limitedcov, "Y")
pc_y_indices_spouses <- bnlearn::spouses(pc_fit_bnlearn_limitedcov, "Y")
print("All relationships identified by PC")
print(pc_y_indices_children)
print(pc_y_indices_parents)
print(pc_y_indices_spouses)

pc_y_indices_all <- unique(c(pc_y_indices_parents, pc_y_indices_children, pc_y_indices_spouses, pc_y_indices))
print("All variables in Markov Blanket of Y identified by PC:")
print(pc_y_indices_all)

# Update your covariate list for the models
cov_pc <- setdiff(pc_y_indices_all, "A") #exclude A
print(cov_pc) #only PAY_0?
```

```{r}

pc_a_indices <- bnlearn::nbr(pc_fit_bnlearn_limitedcov, "A") #all neighbors - but not spouses
print(pc_a_indices)

#parents, children, spouses using cextend version of the graph, which orients some edges and thus allows us to differentiate between parents and children (but still has some undirected edges, so not perfect)
pc_a_indices_parents <- bnlearn::parents(pc_fit_bnlearn_limitedcov, "A")
pc_a_indices_children <- bnlearn::children(pc_fit_bnlearn_limitedcov, "A")
pc_a_indices_spouses <- bnlearn::spouses(pc_fit_bnlearn_limitedcov, "A")
print("All relationships identified by PC")
print(pc_a_indices_children)
print(pc_a_indices_parents)
print(pc_a_indices_spouses)

pc_a_indices_all <- unique(c(pc_a_indices_parents, pc_a_indices_children, pc_a_indices_spouses, pc_a_indices))
print("All variables in Markov Blanket of A identified by PC:")
print(pc_a_indices_all)


cov_pc_a <-setdiff(pc_a_indices_all, "Y") #exclude Y, since we don't want to condition on the outcome in the propensity score model
print("Variables selected by PC (Markov Blanket of A):")
print(cov_pc_a)
```

# pc algorithm nontransformed or
```{r run or on pc}

#outcome regression
if(length(cov_pc) == 0) {
  formula_or_pc <- as.formula("Y ~ A")
} else {
  formula_or_pc <- as.formula(paste("Y ~ A +", paste(cov_pc, collapse = "+")))
}
model_or_pc <- glm(formula_or_pc, data = data, family = binomial)

print("formula")
print(formula_or_pc)

# pc_data_all_enrolled <- data_all_enrolled %>%
#   # mutate(across(everything(), as.numeric)) %>%
#   select(Y, A, all_of(cov_all))

# pc_data_no_enrolled <- data_no_enrolled %>%
#   # mutate(across(everything(), as.numeric)) %>%
#   select(Y, A, all_of(cov_all))

# Predict default probability for world of everyone enrolled, and no one enrolled
data$prob_Y_if_A1_pc_var<- predict(model_or_pc, newdata = data_all_enrolled, type = "response") #type=response gives 0-1 prob

EY_a1_or_pc_var <- mean(data$prob_Y_if_A1_pc_var)

data$prob_Y_if_A0_pc_var <- predict(model_or_pc, newdata = data_no_enrolled, type = "response")
EY_a0_or_pc_var <- mean(data$prob_Y_if_A0_pc_var)

ace_or_pc_var <- EY_a1_or_pc_var - EY_a0_or_pc_var

#print means
print("Default likelihood (PC Set, OR), A=1 then A=0")
print(EY_a1_or_pc_var)
print(EY_a0_or_pc_var)

print(paste("OR ACE (PC Set):", round(ace_or_pc_var,4))) 
```



# PC algorithm nonttransformed ipw

```{r IPW pc var calc ps }

if(length(cov_pc_a) == 0) {
  formula_ps_pc_var <- as.formula("A ~ 1")
} else {
  formula_ps_pc_var <- as.formula(paste("A~", paste(cov_pc_a, collapse = "+"))) #formula of A~pc covariates
}
#get propensity scores
model_ps_pc_var <- glm(formula_ps_pc_var, data = data, family = binomial) #binomial because only 2 outcomes

print("formula")
print(formula_ps_pc_var)

data$ps_pc_var <- predict(model_ps_pc_var, type = "response") #calculate propensity scores for all customers; response transforms to probabilities.
print("min and max ps")
print(min(data$ps_pc_var)) 
print(max(data$ps_pc_var)) 

#extreme ps means likely have extreme scores

data$ps_pc_var_clamped <- pmax(pmin(data$ps_pc_var, 0.9999), 0.0001) 


# calculate weights.
data$weights_pc_var <- ifelse(data[["A"]] == 1, 1 / data$ps_pc_var_clamped, 1 / (1 - data$ps_pc_var_clamped)) 

print("min and max weights")
print(min(data$weights_pc_var)) #1.0001
print(max(data$weights_pc_var)) #2.8


#if customer was actually enrolled (A==1), then d weight as 1/ps, otherwise do 1/(1-ps)
#So if customer had A=1 & ps=0.8, then weight = 1.25; but if A=0 then weight=5. Scales by unexpectedness of actual event.

# Estimate ACE using the weighted average
#filter to only people who were enrolled A=1 or A=0
a0_group <- data[data$A == 0, ]
a1_group <- data[data$A == 1, ]
# Filter to A0 or A1 groups, get default and multiply by weight; basically, if paid on time, contributes nothing to numerator. numerator = total # of defaults, denominator = total # of people (or sum of weights here)
EY_a0_ipw_pc_var <- sum(as.numeric(as.character(a0_group$Y)) * a0_group$weights_pc_var)/sum(a0_group$weights_pc_var)
EY_a1_ipw_pc_var <- sum(as.numeric(as.character(a1_group$Y)) * a1_group$weights_pc_var)/sum(a1_group$weights_pc_var)

ace_ipw_pc_var <- EY_a1_ipw_pc_var - EY_a0_ipw_pc_var
print(paste("IPW ACE (PC set):", round(ace_ipw_pc_var,4)))
```



# Pc algorithm bootstrap nontransformed
```{r bootstrap pc method nontransformed}
bootstrap_results_pc <- replicate(bootstrap_sample_size, {
  #bootstrap sample
  sample_df <- data[sample(nrow(data), replace = TRUE), ]
  
  #convert data to numerical for gaussian tests
  sample_df_pc_data <- sample_df %>%
    mutate(across(where(is.integer), as.numeric)) %>% #convert all integers to numeric, to run
    select(Y, A, all_of(cov_all))

  cov_most_recent <- c("PAY_0", "EDUCATION", "AGE", "SEX", "PAY_AMT1", "BILL_AMT1", "LIMIT_BAL")

  sample_df_pc_data_limitedcov <- sample_df_pc_data %>%
    select(Y, A, all_of(cov_most_recent))

  pc_fit_boot <- pc.stable(sample_df_pc_data_limitedcov, 
                    # blacklist = master_blacklist, 
                      # test = "mix", #not needed, will automatically detect which test based on variable type
                    alpha = 0.01)

  adj_matrix <- bnlearn::amat(pc_fit_boot)

  #extract markov blanket for y
  y_indices_boot <- bnlearn::nbr(pc_fit_boot, "Y") #all neighbors - but not spouses

  #parents, children, spouses using cextend version of the graph, which orients some edges and thus allows us to differentiate between parents and children (but still has some undirected edges, so not perfect)
  y_indices_parents_boot <- bnlearn::parents(pc_fit_boot, "Y")
  y_indices_children_boot <- bnlearn::children(pc_fit_boot, "Y")
  y_indices_spouses_boot <- bnlearn::spouses(pc_fit_boot, "Y")
 
  y_indices_all_boot <- unique(c(y_indices_parents_boot, y_indices_children_boot, y_indices_spouses_boot, y_indices_boot))

  # Update  covariate list for the models
  cov_pc_boot <- setdiff(y_indices_all_boot, "A")


  #now for a
  a_indices_boot <- bnlearn::nbr(pc_fit_boot, "A") #all neighbors - but not spouses
  print(a_indices_boot)

  #parents, children, spouses using cextend version of the graph, which orients some edges and thus allows us to differentiate between parents and children (but still has some undirected edges, so not perfect)
  a_indices_parents_boot <- bnlearn::parents(pc_fit_boot, "A")
  a_indices_children_boot <- bnlearn::children(pc_fit_boot, "A")
  a_indices_spouses_boot <- bnlearn::spouses(pc_fit_boot, "A")

  a_indices_all_boot <- unique(c(a_indices_parents_boot, a_indices_children_boot, a_indices_spouses_boot, a_indices_boot))

  cov_pc_boot_a <-setdiff(a_indices_all_boot, "Y")
  
  # what if the blanket is empty? or
  if(length(cov_pc_boot) == 0) {
    formula_or_pc_boot <- as.formula("Y ~ A")
  } else {
    formula_or_pc_boot <- as.formula(paste("Y ~ A +", paste(cov_pc_boot, collapse = "+")))
  }
  #ipw
  if(length(cov_pc_boot_a) == 0) {
    formula_ps_pc_boot <- as.formula("A ~ 1")
  } else {
    formula_ps_pc_boot <- as.formula(paste("A ~", paste(cov_pc_boot_a, collapse = "+")))
  }
  
  # outcome regression
  model_or_pc_boot <- glm(formula_or_pc_boot, data = sample_df, family = binomial)
  
  #create fake datasets
  sample_df_a0 <- sample_df
  sample_df_a0$A <- factor(0, levels = levels(sample_df$A))
  sample_df_a1 <- sample_df
  sample_df_a1$A <- factor(1, levels = levels(sample_df$A))

  # Predict default probability for world of everyone enrolled, and no one enrolled
  sample_df_prob_Y_if_A1 <- predict(model_or_pc_boot, newdata = sample_df_a1, type = "response") 
  EY_a1_or_sample_df <- mean(sample_df_prob_Y_if_A1)

  sample_df_prob_Y_if_A0 <- predict(model_or_pc_boot, newdata = sample_df_a0, type = "response") 
  EY_a0_or_sample_df <- mean(sample_df_prob_Y_if_A0)
  
  ace_or_sample_df <- EY_a1_or_sample_df - EY_a0_or_sample_df
  
  #IPW
  #get propensity score
  model_ps_sample_df <- glm(formula_ps_pc_boot, data = sample_df, family = binomial)
  sample_df$ps_sample_df <- predict(model_ps_sample_df, type = "response")
  #clamp ps score between 0.0001 and 0.9999 because otherwise get warnings of fitted prob being 0 or 1 occurring
  sample_df$ps_sample_df_clamped <- pmax(pmin(sample_df$ps_sample_df, 0.9999), 0.0001)
  #get weights
  sample_df$weights_sample_df <- ifelse(sample_df$A == 1, 1/sample_df$ps_sample_df_clamped, 1/(1-sample_df$ps_sample_df_clamped))
  #divide into a0 and a1 groups
  a0_group_sample_df <- sample_df[sample_df$A == 0, ]
  a1_group_sample_df <- sample_df[sample_df$A == 1, ]
  #calculate ACE
  EY_a0_ipw_sample_df <- sum(as.numeric(as.character(a0_group_sample_df$Y)) * a0_group_sample_df$weights_sample_df)/sum(a0_group_sample_df$weights_sample_df)
  EY_a1_ipw_sample_df <- sum(as.numeric(as.character(a1_group_sample_df$Y)) * a1_group_sample_df$weights_sample_df)/sum(a1_group_sample_df$weights_sample_df)
  ace_ipw_sample_df <-  EY_a1_ipw_sample_df - EY_a0_ipw_sample_df

  return(list(ace_or = ace_or_sample_df, 
              ace_ipw = ace_ipw_sample_df, 
              adj_matrix = adj_matrix,
              blanket = y_indices_all_boot,
              blanket_a = a_indices_all_boot
              ))
}, simplify = FALSE)
```


```{r bootstrap variance and bias for pc variables}
orig_ace_pc_var <- c(ace_or_pc_var, ace_ipw_pc_var)

#extract ACE values
boot_aces_or_pc  <- unlist(lapply(bootstrap_results_pc, function(x) x$ace_or))
boot_aces_ipw_pc <- unlist(lapply(bootstrap_results_pc, function(x) x$ace_ipw))

#extract markov blanket
boot_blankets_pc <- lapply(bootstrap_results_pc, function(x) x$blanket)
boot_blankets_pc_a <- lapply(bootstrap_results_pc, function(x) x$blanket_a)

# convert to dataframe
boot_df_pc_var <- data.frame(
  Regression = boot_aces_or_pc,
  IPW = boot_aces_ipw_pc
)
colnames(boot_df_pc_var) <- c("Regression", "IPW")

# Calculate Metrics
metrics_pc_var <- data.frame(
  Method = c("Regression", "IPW"),
  Original_ACE = orig_ace_pc_var, # Use values from the full dataset
  Bootstrap_ACE = c(mean(boot_df_pc_var$Regression), mean(boot_df_pc_var$IPW)),
  Standard_Error = c(sd(boot_df_pc_var$Regression), sd(boot_df_pc_var$IPW))
)

# Calculate Bias
metrics_pc_var$Bootstrap_Bias <- metrics_pc_var$Bootstrap_ACE - metrics_pc_var$Original_ACE
print(metrics_pc_var)
```

```{r markov blankets}
boot_blankets_pc_unlisted <- unlist(boot_blankets_pc)

boot_mb_stability_pc <- table(boot_blankets_pc_unlisted) / bootstrap_sample_size
print("Selection Frequency for Y's Markov Blanket:")
print(sort(boot_mb_stability_pc, decreasing = TRUE))
```

```{r markov blankets a}
boot_blankets_pc_a_unlisted <- unlist(boot_blankets_pc_a)

boot_mb_stability_pc_a <- table(boot_blankets_pc_a_unlisted) / bootstrap_sample_size
print("Selection Frequency for A's Markov Blanket:")
print(sort(boot_mb_stability_pc_a, decreasing = TRUE))
```

```{r DAG avg, fig.width=10, fig.height=8, out.width="100%"}
#get adjacent matrices
all_adj_bootstrap_pc <- lapply(bootstrap_results_pc, function(x) x$adj_matrix)
#sum them up to get edge frequencies - prob an edge exists between 2 variables
consensus_matrix_bootstrap_pc <- Reduce("+", all_adj_bootstrap_pc) / bootstrap_sample_size #reduce sums all matrices. if arrow from pay_0 to y is in every bootstrap, then it will have bootstrap_sample_size values in that matrix cell. if only half the time, then half the size.

# Create a graph where an edge exists if it appeared in > 50% of samples
average_adj_bootstrap_pc <- (consensus_matrix_bootstrap_pc > 0.5) * 1

# Convert back to a graph object for plotting
avg_graph_bootstrap_pc <- graph_from_adjacency_matrix(average_adj_bootstrap_pc, mode = "directed")

plot(avg_graph_bootstrap_pc, 
     main = "Consensus DAG (Edges in >50% of Bootstraps): PC Method, Limited Covariates",
     vertex.size = 15,              # Smaller nodes
     vertex.label.cex = 0.6,        # Smaller labels
     vertex.label.dist = 1.2,       # Offset labels from node centers
     edge.arrow.size = 0.3,         # Smaller arrows
     edge.curved = 0.2)             # Slight curve to distinguish overlapping arrows

png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/consensus_dag_pc_bootstrap_limitedcov.png", width = 1200, height = 1000, res = 150)
par(mar = c(1, 1, 3, 1)) 
plot(avg_graph_bootstrap_pc, 
     main = "Consensus DAG (Edges in >50% of Bootstraps): PC Method, Limited Covariates",
     vertex.size = 15,              # Smaller nodes
     vertex.label.cex = 0.6,        # Smaller labels
     vertex.label.dist = 1.2,       # Offset labels from node centers
     edge.arrow.size = 0.3,         # Smaller arrows
     edge.curved = 0.2)             # Slight curve to distinguish overlapping arrows
dev.off()
```



# 5 Now the Analysis, for PC-selected variables! (full cov)

```{r obtain DAG pc full}
pc_data <- data %>%
  mutate(across(where(is.integer), as.numeric)) %>% #convert all integers to numeric, to run
  select(Y, A, all_of(cov_all))
# pc_data_var_types <- ifelse(sapply(pc_data, is.factor), 1, 0)
print(summary(pc_data)) #proves data is in right format.
# attr(pc_data, "adaptDF") <- TRUE

# Run PC; alpha serves as a tuning parameter for the significance of independence tests
pc_fit_test_full_fin <- pc(suffStat = pc_data,
                    indepTest = mixCItest,
                    labels = colnames(pc_data),
                    alpha = 0.01, 
                    verbose = FALSE,
                    u2pd = "relaxed", 
                    skel.method = "stable" ,  #order of test doesnt matter
                    maj.rule = FALSE,
                    solve.confl = TRUE #if some say a-->b and some say b-->a, then pick a direction
                    )

# Plot DAG
plot(pc_fit_test_full_fin@graph, main = "Causal DAG: PC Method, All Covariates")

#export
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_fullcov.png", width = 1200, height = 1000, res = 150)

plot(pc_fit_test_full_fin@graph, main = "Causal DAG: PC Method, All Covariates")

# 3. Close the device to save the file
dev.off()
```
 



```{r obtain markov blanket pc full }

pc_full_dag <- pcalg2dagitty(as(pc_fit_test_full_fin@graph, "matrix"), 
                       labels = colnames(pc_data), 
                       type = "cpdag")

# Get the Markov Blanket (Parents + Children + Spouses)
pc_full_mb_y <- markovBlanket(pc_full_dag, "Y")
print(pc_full_mb_y)

# Update your covariate list for the models
cov_pc_full <- setdiff(pc_full_mb_y, "A") #exclude A
print(cov_pc_full) #only PAY_0?
```

```{r}
# Get the Markov Blanket (Parents + Children + Spouses)
pc_full_mb_a <- markovBlanket(pc_full_dag, "A")
print(pc_full_mb_a)

# Update your covariate list for the models
cov_pc_full_a <- setdiff(pc_full_mb_a, "Y") #exclude A
print(cov_pc_full_a) #only PAY_0?
```

# pc algorithm nontransformed or
```{r run or on pc full}

#outcome regression
if(length(cov_pc_full) == 0) {
  formula_or_pc_full <- as.formula("Y ~ A")
} else {
  formula_or_pc_full <- as.formula(paste("Y ~ A +", paste(cov_pc_full, collapse = "+")))
}
model_or_pc_full <- glm(formula_or_pc_full, data = data, family = binomial)

print("formula")
print(formula_or_pc_full)

# pc_data_all_enrolled <- data_all_enrolled %>%
#   # mutate(across(everything(), as.numeric)) %>%
#   select(Y, A, all_of(cov_all))

# pc_data_no_enrolled <- data_no_enrolled %>%
#   # mutate(across(everything(), as.numeric)) %>%
#   select(Y, A, all_of(cov_all))

# Predict default probability for world of everyone enrolled, and no one enrolled
data$prob_Y_if_A1_pc_var_full<- predict(model_or_pc_full, newdata = data_all_enrolled, type = "response") #type=response gives 0-1 prob

EY_a1_or_pc_var_full <- mean(data$prob_Y_if_A1_pc_var_full)

data$prob_Y_if_A0_pc_var_full <- predict(model_or_pc_full, newdata = data_no_enrolled, type = "response")
EY_a0_or_pc_var_full <- mean(data$prob_Y_if_A0_pc_var_full)

ace_or_pc_var_full <- EY_a1_or_pc_var_full - EY_a0_or_pc_var_full

#print means
print("Default likelihood (PC Set, OR), A=1 then A=0")
print(EY_a1_or_pc_var_full)
print(EY_a0_or_pc_var_full)

print(paste("OR ACE (PC Set):", round(ace_or_pc_var_full,4))) 
```



# PC algorithm nonttransformed ipw

```{r IPW pc var calc ps full}

if(length(cov_pc_full_a) == 0) {
  formula_ps_pc_var_full <- as.formula("A ~ 1")
} else {
  formula_ps_pc_var_full <- as.formula(paste("A~", paste(cov_pc_full_a, collapse = "+"))) #formula of A~pc covariates
}
#get propensity scores
model_ps_pc_var_full <- glm(formula_ps_pc_var_full, data = data, family = binomial) #binomial because only 2 outcomes

print("formula")
print(formula_ps_pc_var_full)

data$ps_pc_var_full <- predict(model_ps_pc_var_full, type = "response") #calculate propensity scores for all customers; response transforms to probabilities.
print("min and max ps")
print(min(data$ps_pc_var_full)) 
print(max(data$ps_pc_var_full)) 

#extreme ps means likely have extreme scores

data$ps_pc_var_clamped_full <- pmax(pmin(data$ps_pc_var_full, 0.9999), 0.0001) 


# calculate weights.
data$weights_pc_var_full <- ifelse(data[["A"]] == 1, 1 / data$ps_pc_var_clamped_full, 1 / (1 - data$ps_pc_var_clamped_full)) 

print("min and max weights")
print(min(data$weights_pc_var_full)) #1.0001
print(max(data$weights_pc_var_full)) #2.8

#if customer was actually enrolled (A==1), then d weight as 1/ps, otherwise do 1/(1-ps)
#So if customer had A=1 & ps=0.8, then weight = 1.25; but if A=0 then weight=5. Scales by unexpectedness of actual event.

# Estimate ACE using the weighted average
#filter to only people who were enrolled A=1 or A=0
a0_group <- data[data$A == 0, ]
a1_group <- data[data$A == 1, ]
# Filter to A0 or A1 groups, get default and multiply by weight; basically, if paid on time, contributes nothing to numerator. numerator = total # of defaults, denominator = total # of people (or sum of weights here)
EY_a0_ipw_pc_var_full <- sum(as.numeric(as.character(a0_group$Y)) * a0_group$weights_pc_var_full)/sum(a0_group$weights_pc_var_full)
EY_a1_ipw_pc_var_full <- sum(as.numeric(as.character(a1_group$Y)) * a1_group$weights_pc_var_full)/sum(a1_group$weights_pc_var_full)

ace_ipw_pc_var_full <- EY_a1_ipw_pc_var_full - EY_a0_ipw_pc_var_full
print(paste("IPW ACE (PC set):", round(ace_ipw_pc_var_full,4)))
```




# 6 Now the Analysis, for Local discovery methods

Use MMPC from bnlearn. The max-min parents and children algorithm. This searches the neighborhood of Y (parents and children) by testing for conditional independence at the alpha=0.01 level. Variable is only kept if relationship with Y can't be explained away by other variables.

# nontransformed, local method

IAMB (Incremental Association)

The most common "Gold Standard" for medium-sized datasets.

    How it works: It is dynamic. In every step of the growth phase, it re-tests all remaining variables and picks the one with the highest Mutual Information (or lowest p-value) given the current blanket.

    The Pros: Much more reliable at finding the correct parents, children, and spouses than GS.

    The Cons: Slower than GS (it runs O(n2) tests).

    N=500 Performance: Recommended. This is the most balanced choice for your credit data.

# Test IAMB

```{r}
#convert data to numerical 
local_data_test <- data %>%
  mutate(across(where(is.integer), as.numeric)) %>% #convert all integers to numeric, to run
  select(Y, A, all_of(cov_all))

local_data_test_mbonly <- local_data_test %>%
    mutate(across(everything(), function(x) {
      if(is.factor(x) || is.character(x)) as.numeric(as.factor(x)) else rank(x)
    }))

# Run nbr to find the markov blanket of Y only
local_y_test <- learn.mb(local_data_test_mbonly, node = "Y", method = "iamb", alpha = 0.01, test="cor")

print("Variables in the Local Neighborhood (Parents/Children) of Y:")
print(local_y_test)

# Extract adjustment set (exclude treatment A)
cov_local_test <- setdiff(local_y_test, "A") #L
print(cov_local_test)


#now do the same for A to get covariates for propensity score model
local_a_test <- learn.mb(local_data_test_mbonly, node = "A", method = "iamb", alpha = 0.01, test="cor")
print("Variables in the Local Neighborhood (Parents/Children) of A:")
print(local_a_test)
cov_local_a_test <-setdiff(local_a_test, "Y") #exclude Y, since we don't want to condition on the outcome in the propensity score model
print(cov_local_a_test)
```

```{r}
# learn.mb(local_data, node = "A", method = "a", alpha = 0.01)
#Error in check.label(algorithm, choices = ok, labels = learning.labels,  :  valid learning algorithm(s) are "gs" (Grow-Shrink), "iamb" (IAMB), "fast.iamb" (Fast-IAMB), "inter.iamb" (Inter-IAMB), "iamb.fdr" (IAMB-FDR). See ?bnlearn-package for details
```


```{r}
#convert data to numerical 
local_data <- data %>%
  mutate(across(where(is.integer), as.numeric)) %>% #convert all integers to numeric, to run
  select(Y, A, all_of(cov_all))

local_data_test_mbonly <- local_data %>%
    mutate(across(everything(), function(x) {
      if(is.factor(x) || is.character(x)) as.numeric(as.factor(x)) else rank(x)
    }))


# Run nbr to find the markov blanket of Y only
local_y <- learn.mb(local_data_test_mbonly, node = "Y", method = "inter.iamb", alpha = 0.01, test = "cor")

print("Variables in the Local Neighborhood (Parents/Children) of Y:")
print(local_y)

# Extract adjustment set (exclude treatment A)
cov_local <- setdiff(local_y, "A") #L
print(cov_local)


#now do the same for A to get covariates for propensity score model
local_a <- learn.mb(local_data_test_mbonly, node = "A", method = "inter.iamb", alpha = 0.01, test = "cor")
print("Variables in the Local Neighborhood (Parents/Children) of A:")
print(local_a)
cov_local_a <-setdiff(local_a, "Y") #exclude Y, since we don't want to condition on the outcome in the propensity score model
print(cov_local_a)
```


# local algorithm nontransformed or

```{r or local algorithm}
# --- Outcome Regression (Local Set) ---
if (length(cov_local) == 0) {
  formula_or_local <- as.formula("Y ~ A")
} else {
  formula_or_local <- as.formula(paste("Y ~ A +", paste(cov_local, collapse = " + ")))
}

model_or_local <- glm(formula_or_local, data = data, family = binomial)

# Counterfactual predictions
EY_a1_or_local <- mean(predict(model_or_local, newdata = data_all_enrolled, type = "response"))
EY_a0_or_local <- mean(predict(model_or_local, newdata = data_no_enrolled, type = "response"))
ace_or_local <- EY_a1_or_local - EY_a0_or_local

print(paste("Local Discovery OR ACE:", round(ace_or_local, 4)))
```

# local algorithm nontransformed ipw

```{r ipw local algorithm}
if (length(cov_local_a) == 0) {
  formula_ps_local <- as.formula("A ~ 1")
} else {
  formula_ps_local <- as.formula(paste("A ~", paste(cov_local_a, collapse = " + ")))
}

model_ps_local <- glm(formula_ps_local, data = data, family = binomial)
data$ps_local <- predict(model_ps_local, type = "response")
data$ps_local_clamped <- pmax(pmin(data$ps_local, 0.9999), 0.0001)

# Weights
data$weights_local <- ifelse(data$A == 1, 1 / data$ps_local_clamped, 1 / (1 - data$ps_local_clamped))

a0_group <- data[data$A == 0, ]
a1_group <- data[data$A == 1, ]


EY_a0_ipw_local <- sum(as.numeric(as.character(a0_group$Y)) * a0_group$weights_local)/sum(a0_group$weights_local)
EY_a1_ipw_local <- sum(as.numeric(as.character(a1_group$Y)) * a1_group$weights_local)/sum(a1_group$weights_local)

ace_ipw_local <- EY_a1_ipw_local - EY_a0_ipw_local
print(paste("Local Discovery IPW ACE:", round(ace_ipw_local, 4)))
```


# local algorithm nontransformed bootstrap

```{r bootstrap local}
bootstrap_results_local <- replicate(bootstrap_sample_size, {
  sample_df <- data[sample(nrow(data), replace = TRUE), ]
  

  sample_df_local <- sample_df %>%
    mutate(across(where(is.integer), as.numeric)) %>% #convert all integers to numeric, to run
    select(Y, A, all_of(cov_all))

  # This maps factors to integers and continuous variables to ranks
  sample_df_ranked <- sample_df_local %>%
    mutate(across(everything(), function(x) {
      if(is.factor(x) || is.character(x)) as.numeric(as.factor(x)) else rank(x)
    }))

  # We use sample_df_ranked, method = "inter.iamb", and test = "cor"
  local_y_boot <- tryCatch({
    learn.mb(sample_df_ranked, node = "Y", method = "inter.iamb", test = "cor", alpha = 0.01)
  }, error = function(e) character(0))
  
  cov_local_boot <- setdiff(local_y_boot, "A")
  
  local_a_boot <- tryCatch({
    learn.mb(sample_df_ranked, node = "A", method = "inter.iamb", test = "cor", alpha = 0.01)
  }, error = function(e) character(0))
  
  cov_local_boot_a <- setdiff(local_a_boot, "Y")
  
 
  if (length(cov_local_boot) == 0) {
    formula_or_local_boot <- as.formula("Y ~ A")
  } else {
    formula_or_local_boot <- as.formula(paste("Y ~ A +", paste(cov_local_boot, collapse = " + ")))
  }
  if (length(cov_local_boot_a) == 0) {
    formula_ps_local_boot <- as.formula("A ~ 1")
  } else {
    formula_ps_local_boot <- as.formula(paste("A ~", paste(cov_local_boot_a, collapse = " + ")))
  }


  # Uses sample_df so the GLM correctly processes the original factors/scales
  model_or_local_boot <- glm(formula_or_local_boot, data = sample_df, family = binomial)
  
  # create fake datasets
  sample_df_a0 <- sample_df
  sample_df_a0$A <- factor(0, levels = levels(sample_df$A))
  sample_df_a1 <- sample_df
  sample_df_a1$A <- factor(1, levels = levels(sample_df$A))

  sample_df_prob_Y_if_A1 <- predict(model_or_local_boot, newdata = sample_df_a1, type = "response")
  EY_a1_or_sample_df <- mean(sample_df_prob_Y_if_A1)

  sample_df_prob_Y_if_A0 <- predict(model_or_local_boot, newdata = sample_df_a0, type = "response")
  EY_a0_or_sample_df <- mean(sample_df_prob_Y_if_A0)
  
  ace_or_sample_df <- EY_a1_or_sample_df - EY_a0_or_sample_df
  
  # IPW
  model_ps_local_boot <- glm(formula_ps_local_boot, data = sample_df, family = binomial)
  sample_df$ps_sample_df <- predict(model_ps_local_boot, type = "response")
  sample_df$ps_sample_df_clamped <- pmax(pmin(sample_df$ps_sample_df, 0.9999), 0.0001)
  sample_df$weights_sample_df <- ifelse(sample_df$A == 1, 1/sample_df$ps_sample_df_clamped, 1/(1-sample_df$ps_sample_df_clamped))

  a0_group_sample_df <- sample_df[sample_df$A== 0, ]
  a1_group_sample_df <- sample_df[sample_df$A== 1, ]

  EY_a0_ipw_sample_df <- sum(as.numeric(as.character(a0_group_sample_df$Y)) * a0_group_sample_df$weights_sample_df) / sum(a0_group_sample_df$weights_sample_df)
  EY_a1_ipw_sample_df <- sum(as.numeric(as.character(a1_group_sample_df$Y)) * a1_group_sample_df$weights_sample_df) / sum(a1_group_sample_df$weights_sample_df) 
  
  ace_ipw_sample_df <- EY_a1_ipw_sample_df - EY_a0_ipw_sample_df
  
  return(list(ace_or = ace_or_sample_df, 
       ace_ipw = ace_ipw_sample_df, 
       blanket = local_y_boot,
       blanket_a = local_a_boot
       ))
}, simplify=FALSE)

```



```{r bootstrap variance and bias for local variables}
orig_ace_local_var <- c(ace_or_local, ace_ipw_local)

#extract ACE values
boot_aces_or_local  <- unlist(lapply(bootstrap_results_local, function(x) x$ace_or))
boot_aces_ipw_local <- unlist(lapply(bootstrap_results_local, function(x) x$ace_ipw))

#extract markov blanket
boot_blankets_local <- lapply(bootstrap_results_local, function(x) x$blanket)
boot_blankets_local_a <- lapply(bootstrap_results_local, function(x) x$blanket_a)


# convert to dataframe
boot_df_local_var <- data.frame(
  Regression = boot_aces_or_local,
  IPW = boot_aces_ipw_local
)
colnames(boot_df_local_var) <- c("Regression", "IPW")

# Calculate Metrics
metrics_local_var <- data.frame(
  Method = c("Regression", "IPW"),
  Original_ACE = orig_ace_local_var,
  Bootstrap_ACE = c(mean(boot_df_local_var$Regression), mean(boot_df_local_var$IPW)),
  Standard_Error = c(sd(boot_df_local_var$Regression), sd(boot_df_local_var$IPW))
)

# Calculate Bias
metrics_local_var$Bootstrap_Bias <- metrics_local_var$Bootstrap_ACE - metrics_local_var$Original_ACE
print(metrics_local_var)
```

```{r}
boot_blankets_local_unlisted <- unlist(boot_blankets_local)

boot_mb_stability_local <- table(boot_blankets_local_unlisted) / bootstrap_sample_size
print("Selection Frequency for Y's Markov Blanket:")
print(sort(boot_mb_stability_local, decreasing = TRUE))
```

```{r}
boot_blankets_local_unlisted_a <- unlist(boot_blankets_local_a)

boot_mb_stability_local_a <- table(boot_blankets_local_unlisted_a) / bootstrap_sample_size
print("Selection Frequency for A's Markov Blanket:")
print(sort(boot_mb_stability_local_a, decreasing = TRUE))
```


# 7 Now for the iamb with blacklist
```{r create blacklist local nontransformed forbidden edges}

allowed_nodes_local_forbidden_edges <- c("Y", "A", cov_all)
# Nothing can cause Demographics
bl_demog <- expand.grid(from = allowed_nodes_local_forbidden_edges, to = c("AGE", "SEX", "EDUCATION"))

# Y cannot cause anything
bl_y <- expand.grid(from = "Y", to = setdiff(allowed_nodes_local_forbidden_edges, "Y"))

# A cannot cause anything except Y
bl_a <- expand.grid(from = "A", to = setdiff(allowed_nodes_local_forbidden_edges, c("A", "Y")))

# temporal relationships: later months cannot cause earlier months
pay_vars <- c("PAY_0", "PAY_2", "PAY_3", "PAY_4", "PAY_5", "PAY_6")
bill_amt_vars <- c("BILL_AMT1", "BILL_AMT2", "BILL_AMT3", "BILL_AMT4", "BILL_AMT5", "BILL_AMT6")
pay_amt_vars <- c("PAY_AMT1", "PAY_AMT2", "PAY_AMT3", "PAY_AMT4", "PAY_AMT5", "PAY_AMT6")

bl_temporal <- data.frame()
for(i in 1:(length(pay_vars)-1)) {
  for(j in (i+1):length(pay_vars)) {
    # Block Present -> Past
    bl_temporal <- rbind(bl_temporal, data.frame(from = pay_vars[i], to = pay_vars[j]))
    bl_temporal <- rbind(bl_temporal, data.frame(from = bill_amt_vars[i], to = bill_amt_vars[j]))
    bl_temporal <- rbind(bl_temporal, data.frame(from = pay_amt_vars[i], to = pay_amt_vars[j]))
  }
}

# Combine all into one master blacklist
master_blacklist <- rbind(bl_demog, bl_y, bl_a, bl_temporal) %>%
  mutate(across(everything(), as.character)) %>% # Convert everything to character to avoid factor level mismatch errors
  filter(from != to) %>% # Remove self-loops
  distinct()             # Remove duplicates

print("Master Blacklist:")
print(master_blacklist)
```

# nontransformed, local method

IAMB (Incremental Association)

The most common "Gold Standard" for medium-sized datasets.

    How it works: It is dynamic. In every step of the growth phase, it re-tests all remaining variables and picks the one with the highest Mutual Information (or lowest p-value) given the current blanket.

    The Pros: Much more reliable at finding the correct parents, children, and spouses than GS.

    The Cons: Slower than GS (it runs O(n2) tests).

    N=500 Performance: Recommended. This is the most balanced choice for your credit data.


```{r}
#convert data to numerical 
local_data <- data %>%
  mutate(across(where(is.integer), as.numeric)) %>% #convert all integers to numeric, to run
  select(Y, A, all_of(cov_all))

#because local discovery, need to exclude Y from blacklist...
master_blacklist_excl_y <- master_blacklist %>%
  filter(from != "Y" & to != "Y") %>%
  as.matrix()

master_blacklist_matrix_excl_y <- as.matrix(master_blacklist_excl_y)

# Run nbr to find the markov blanket of Y only
local_y_blacklist <- learn.mb(local_data, node = "Y", method = "inter.iamb", blacklist = master_blacklist_matrix_excl_y, alpha = 0.01)

print("Variables in the Local Neighborhood (Parents/Children) of Y:")
print(local_y_blacklist)

# Extract adjustment set (exclude treatment A)
cov_local_blacklist <- setdiff(local_y_blacklist, "A") #L
print(cov_local_blacklist)


master_blacklist_excl_a <- master_blacklist %>%
  filter(from != "A" & to != "A") %>%
  as.matrix()

master_blacklist_matrix_excl_a <- as.matrix(master_blacklist_excl_a)


#now do the same for A to get covariates for propensity score model
local_a_blacklist <- learn.mb(local_data, node = "A", method = "inter.iamb", blacklist = master_blacklist_matrix_excl_a, alpha = 0.01)
print("Variables in the Local Neighborhood (Parents/Children) of A:")
print(local_a_blacklist)
cov_local_a_blacklist <-setdiff(local_a_blacklist, "Y") #exclude Y, since we don't want to condition on the outcome in the propensity score model
print(cov_local_a_blacklist)
```


# local algorithm nontransformed or

```{r or local algorithm blacklist}
# --- Outcome Regression (Local Set) ---
if (length(cov_local_blacklist) == 0) {
  formula_or_local_blacklist <- as.formula("Y ~ A")
} else {
  formula_or_local_blacklist <- as.formula(paste("Y ~ A +", paste(cov_local_blacklist, collapse = " + ")))
}

model_or_local_blacklist <- glm(formula_or_local_blacklist, data = data, family = binomial)

# Counterfactual predictions
EY_a1_or_local_blacklist <- mean(predict(model_or_local_blacklist, newdata = data_all_enrolled, type = "response"))
EY_a0_or_local_blacklist <- mean(predict(model_or_local_blacklist, newdata = data_no_enrolled, type = "response"))
ace_or_local_blacklist <- EY_a1_or_local_blacklist - EY_a0_or_local_blacklist

print(paste("Local Discovery OR ACE:", round(ace_or_local_blacklist, 4)))
```

# local algorithm nontransformed ipw

```{r ipw local algorithm blacklist}
if (length(cov_local_a_blacklist) == 0) {
  formula_ps_local_blacklist <- as.formula("A ~ 1")
} else {
  formula_ps_local_blacklist <- as.formula(paste("A ~", paste(cov_local_a_blacklist, collapse = " + ")))
}

model_ps_local_blacklist <- glm(formula_ps_local_blacklist, data = data, family = binomial)
data$ps_local_blacklist <- predict(model_ps_local_blacklist, type = "response")
data$ps_local_clamped_blacklist <- pmax(pmin(data$ps_local_blacklist, 0.9999), 0.0001)

# Weights
data$weights_local_blacklist <- ifelse(data$A == 1, 1 / data$ps_local_clamped_blacklist, 1 / (1 - data$ps_local_clamped_blacklist))

a0_group <- data[data$A == 0, ]
a1_group <- data[data$A == 1, ]


EY_a0_ipw_local_blacklist <- sum(as.numeric(as.character(a0_group$Y)) * a0_group$weights_local_blacklist)/sum(a0_group$weights_local_blacklist)
EY_a1_ipw_local_blacklist <- sum(as.numeric(as.character(a1_group$Y)) * a1_group$weights_local_blacklist)/sum(a1_group$weights_local_blacklist)

ace_ipw_local_blacklist <- EY_a1_ipw_local_blacklist - EY_a0_ipw_local_blacklist
print(paste("Local Discovery IPW ACE:", round(ace_ipw_local_blacklist, 4)))
```


```{r bootstrap variance and bias for local variables blacklist}
orig_ace_local_var_blacklist <- c(ace_or_local_blacklist, ace_ipw_local_blacklist)
```


# 8 pc with only most recent month bnlearn, with blacklist


```{r obtain DAG pc blacklist}
#convert data to numerical for gaussian tests
pc_data <- data %>%
  mutate(across(where(is.integer), as.numeric)) %>% #convert all integers to numeric, to run
  select(Y, A, all_of(cov_all))

# calculate correlation matrix and sample size; use to test for conditional independence amongst pairs
# pc_corr_size <- list(C = cor(pc_data), n = nrow(pc_data))


pc_fit_bnlearn_blacklist <- pc.stable(pc_data, 
                    blacklist = master_blacklist, 
                      # test = "mix", #not needed, will automatically detect which test based on variable type
                    alpha = 0.01)


# Get adjacency matrix from PC fit
# pc_fit_adj_matrix <- amat(pc_fit)
# print(pc_fit_adj_matrix)

pc_fit_bnlearn_cextend_blacklist <- cextend(pc_fit_bnlearn_blacklist) #directs edges as much as possible, but doesn't orient all edges. Still has some undirected edges.


pc_fit_bnlearn_maxsx_blacklist <- pc.stable(pc_data, 
                    blacklist = master_blacklist, 
                      # test = "mix", #not needed, will automatically detect which test based on variable type
                    alpha = 0.01,
                    max.sx=2) #limits size of conditioning set to 2, which should help with the power issue of mixCItest, but also means we won't find as many independencies and thus will have a more connected graph. 
pc_fit_bnlearn_maxsx_cextend_blacklist <- cextend(pc_fit_bnlearn_maxsx_blacklist) #directs edges as much as possible, but doesn't orient all edges. Still has some undirected edges.
```


```{r}

# Plot DAG
graphviz.plot(pc_fit_bnlearn_blacklist, shape = "ellipse",
     main = "TEST Full Causal DAG, PC Method")

# 2. Export a high-quality version
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_bnlearn_blacklist_test.png", width = 1600, height = 1200, res = 150)
graphviz.plot(pc_fit_bnlearn_blacklist, shape = "ellipse", 
     main = "TEST Full Causal DAG, PC Method")
# 3. Close the device to save the file
dev.off()
```

```{r}
# Plot DAG
graphviz.plot(pc_fit_bnlearn_cextend_blacklist, shape = "ellipse",
     main = "TEST Full Causal DAG, PC Method")

# 2. Export a high-quality version
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_bnlearn_cextend_blacklist_test.png", width = 1600, height = 1200, res = 150)
graphviz.plot(pc_fit_bnlearn_cextend_blacklist, shape = "ellipse", 
     main = "TEST Full Causal DAG, PC Method")
# 3. Close the device to save the file
dev.off()

```

```{r}

# Plot DAG
graphviz.plot(pc_fit_bnlearn_maxsx_blacklist, shape = "ellipse",
     main = "TESTFull Causal DAG, PC Method")

# 2. Export a high-quality version
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_bnlearn_maxsx_blacklist_test.png", width = 1600, height = 1200, res = 150)
graphviz.plot(pc_fit_bnlearn_maxsx_blacklist, shape = "ellipse", 
     main = "TEST Full Causal DAG, PC Method")
# 3. Close the device to save the file
dev.off()
```

```{r}
# Plot DAG
graphviz.plot(pc_fit_bnlearn_maxsx_cextend_blacklist, shape = "ellipse",
     main = "TEST Full Causal DAG, PC Method")

# 2. Export a high-quality version
png("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/causal_dag_pc_method_bnlearn_cextend_maxsx_blacklist_test.png", width = 1600, height = 1200, res = 150)
graphviz.plot(pc_fit_bnlearn_maxsx_cextend_blacklist, shape = "ellipse", 
     main = "TEST Full Causal DAG, PC Method")
# 3. Close the device to save the file
dev.off()

```

This doesn't work very well, no point in blacklisting...


# 9 end Note: compare the ACEs for all of the runs at the end!! maybe do a big table to discuss.

```{r final ACE comparisons}
# Define a helper function to label and combine
combine_metrics <- function(df, selection_name, data_type) {
  df %>%  
    mutate(
      Selection = selection_name,
      Rel_Bias_Pct = ((Bootstrap_ACE - Original_ACE) / Original_ACE) * 100
      # Bias_SE_Ratio = abs(Bootstrap_ACE - Original_ACE) / Standard_Error
    ) %>%
    mutate(across(where(is.numeric), ~round(.x, 4))) %>%
    mutate(Rel_Bias_Pct = round(Rel_Bias_Pct, 2)) %>%
    select(Selection, Method, Original_ACE, Bootstrap_ACE, everything())
}

# Create a master table
all_metrics_master <- bind_rows(
  combine_metrics(metrics_all_var, "All Variables"),
  combine_metrics(metrics_lasso, "Lasso"),
  combine_metrics(metrics_pc_var, "PC, Limited Covariates",),
  combine_metrics(metrics_rank_pc_var, "Rank PC"),
  combine_metrics(metrics_local_var, "IAMB"),
  # combine_metrics(metrics_pc_var_full, "PC, All Covariates")
  # combine_metrics(metrics_local_var_blacklist, "IAMB with Blacklist")
)

print(all_metrics_master)
```

```{r}

# 1. Prepare the data (Rounding and Renaming)
table_data <- all_metrics_master %>%
  mutate(
    # Match the decimal logic from your gt code
    Original_ACE = round(Original_ACE, 4),
    Bootstrap_ACE = round(Bootstrap_ACE, 4),
    Standard_Error = round(Standard_Error, 4),
    Bootstrap_Bias = round(Bootstrap_Bias, 4),
    Rel_Bias_Pct = paste0(round(Rel_Bias_Pct, 1), "%") # Format as string for ggplot
    # Bias_SE_Ratio = round(Bias_SE_Ratio, 3)
  ) %>%
  # Rename columns to match your desired header
  rename(
    `Estimation Method` = Method,
    `Original ACE` = Original_ACE,
    `Bootstrap Mean ACE` = Bootstrap_ACE,
    `Bootstrap Standard Error (SE)` = Standard_Error,
    `Bootstrap Bias` = Bootstrap_Bias,
    `Relative Bootstrap Bias (%)` = Rel_Bias_Pct,
    # `Bootstrap Bias-to-SE Ratio` = Bias_SE_Ratio
  )

# 2. Create the table graphic
# We use 'ttheme_minimal' or 'ttheme_clean' for a thesis look
tab_plot <- ggtexttable(table_data, 
                        rows = NULL, 
                        theme = ttheme("classic", base_size = 10)) %>%
            tab_add_title(text = "ACE Estimation Comparison Across Variable Selection Methods", 
                          face = "bold", padding = unit(1.5, "line"))

# 3. Save as PNG using ggsave (No Chrome needed!)
ggsave(filename = "OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/ACE_Comparison_Table.png", 
       plot = tab_plot, 
       width = 12, height = 8, dpi = 300, bg = "white")

# View it in RStudio
print(tab_plot)

# # 1. Create the professional gt table object
# final_table_img <- all_metrics_master %>% 
#   gt(groupname_col = "Selection") %>% # Group by Selection method
  
#   # Add title and subtitle for your thesis
#   tab_header(
#     title = md("ACE Estimation Comparison Across Variable Selection Methods"),
#   ) %>%
  
#   # Format column labels clearly
#   cols_label(
#     Method = md("Estimation Method"),
#     Original_ACE = md("Original ACE"),
#     Bootstrap_ACE = md("Bootstrap Mean ACE"),
#     Standard_Error = md("Bootstrap Standard Error (SE)"),
#     Rel_Bias_Pct = md("Relative Bootstrap Bias (%)"),
#     Bias_SE_Ratio = md("Bootstrap Bias-to-SE Ratio"),
#     Bootstrap_Bias = md("Bootstrap Bias")
#   ) %>%
  
#   # Add formatting for scientific precision and percentages
#   fmt_number(columns = c(Original_ACE, Bootstrap_ACE, Standard_Error), decimals = 4) %>%
#   fmt_percent(columns = Rel_Bias_Pct, decimals = 1, scale_values=FALSE) %>%
#   fmt_number(columns = Bias_SE_Ratio, decimals = 3) %>%
  
#   # Choose a clean theme (like minimal or pff)
#   gt_theme_pff()

# # 2. Save the table as an image (PNG)
# # Ensure webshot2 is installed for this step
# gtsave(final_table_img, filename = "OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/ACE_Comparison_Table.png")

# # Print the table to the Viewer to see it immediately
# final_table_img
```

```{r final model comparisons}
print("Formula for ordinary regression model for all variables:")
print(formula_or_all_var)
print("Formula for IPW model for all variables:")
print(formula_ps_all_var)

print("Formula for ordinary regression model for lasso:")
print("It gets cut off so I just output the covariates instead of a full formula as well")
print(formula_lasso_or_text)
print(formula_lasso_or_text_anotherver)
print("Formula for IPW model for lasso variable selection:")
print("It gets cut off so I just output the covariates instead of a full formula as well")
print(formula_lasso_ps_text)
print(formula_lasso_ps_text_anotherver)

print("Formula for ordinary regression model for PC, limiting to most recent month:")
print(formula_or_pc)
print("Formula for IPW for PC, limiting to most recent month:")
print(formula_ps_pc_var)

print("Formula for ordinary regression model for Rank PC:")
print(formula_or_rank_pc)
print("Formula for IPW for Rank PC:")
print(formula_ps_rank_pc_var)

print("Formula for ordinary regression model for local discovery (INTER_IAMB)")
print(formula_or_local)
print("formula for IPW for local discovery (INTER_IAMB)")
print(formula_ps_local)

# print("Formula for ordinary regression model for PC with ALL covariates")
# print(formula_or_pc_full)
# print("formula for IPW for PC with ALL covariates")
# print(formula_ps_pc_var_full)
```


```{r plot all ACEs}


# Calculate 95% Confidence Intervals using the Normal Approximation: ACE +/- 1.96 * SE
all_metrics_master <- all_metrics_master %>%
 mutate(
    CI_lower = Bootstrap_ACE - (1.96 * Standard_Error),
    CI_upper = Bootstrap_ACE + (1.96 * Standard_Error)
  )
ace_se_graph <- ggplot(all_metrics_master, aes(x = Selection, y = Standard_Error, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge") + 
  scale_fill_manual(values = c("Regression" = "#2c7fb8", "IPW" = "#feb24c")) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Comparison of Model Variance (Standard Error)",
       y = "Standard Error of ACE", x = "Variable Selection Method")+
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),    # Slightly smaller text to save space
    legend.title = element_text(size = 9, face = "bold"),
    legend.key.size = unit(0.4, "cm")        # Smaller keys to fit more on screen
  )
ggsave("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/ace_standard_errors.png", plot = ace_se_graph, width = 8, height = 6, dpi = 300)
print(ace_se_graph)

ace_ci_graph <- ggplot(all_metrics_master, aes(x = Selection, y = Bootstrap_ACE, color = Method, group = Method)) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                position = position_dodge(width = 0.9), 
                width = 0.5, alpha = 0.6) +
  geom_point(aes(y = Bootstrap_ACE), 
             position = position_dodge(width = 0.9), 
             size = 3) +
  geom_point(aes(y = Original_ACE), 
             shape = 4, 
             position = position_dodge(width = 0.9), 
             size = 4, stroke = 1.2) +
  coord_flip() +
  theme_minimal() +
  scale_color_manual(values = c("Regression" = "#2c7fb8", "IPW" = "#feb24c")) +
  labs(title = "Average Causal Effect (ACE) Comparisons",
       subtitle = "Dots (•) = Bootstrap Mean ACE (w/ 95% CI) | Crosses (x) = Original Estimate of ACE",
       y = "Estimated ACE (Effect on Default Probability)",
       x = "Variable Selection Strategy",
       color = "Method",
       shape = "Data Format") +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_line(color = "grey90") # Makes the rows easier to follow
  )

ggsave("OUTPUT_IMAGES_FROM_ANALYSIS_V6_01/ace_confidence_intervals.png", plot = ace_ci_graph, width = 8, height = 6, dpi = 300)
print(ace_ci_graph)
```
