---
title: "Project2Submission"
output: pdf_document
---

# Nikita's Notes (outside of results)
- Data was cleaned to remove ::: nonincremental

- Walking only has data for sessions 1 & 3

- Standing does not have data for session 2 on day 2

- Explain principle of marginality in statistical methods?

- How to justify use of LMM - discuss more...

- Add coefficients for each equation, and discuss whether used ML or REML (and impacts?)

- Explain ICC 

- When discussing ICCs, just create tables to show it?? Each row is a model, and each column is a variable?

- Add final summary table to show coefficients, etc of final model; add discussion!

- discuss adding activity to ar1 - better correlation? 

```{r display settings, echo = TRUE, warning = FALSE, message = FALSE, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, include=TRUE)
# echo = FALSE: do not display code
# warning/message = FALSE: do not display those
# include = FALSE: do not display any of the outputs from a code chunk (this is overriden for select chunks we want displayed in the report later)
#to run with code & all chunks shown, set echo = TRUE and include = TRUE here
```

```{r packages}
#download packages
library(tidyverse) #installed tidyverse
library(nlme)
library(ggplot2)
library(here)
library(kableExtra)
library(modelsummary)
library(dplyr)
library(performance)
```

```{r load data}

file_path <- here("analysis_and_plots/heat_flow_ICE_EPFL.csv")

#Load the dataset
data <- read.csv(file_path)

#check data
head(data)
colSums(is.na(data)) #98 NA values
nrow(data)
```

```{r check missing data}
#are NAs by sex?
sex_na_summary <- data %>%
  group_by(sex) %>%
  summarise(
    Total_Rows = n(),
    Missing_HF = sum(is.na(HF_value)),
    Percent_Missing = round((Missing_HF / Total_Rows) * 100, 2)
  )
print(sex_na_summary)

#are NAs by activity and session and day?
activity_session_day_na_summary <- data %>%
  group_by(activity, session, day) %>%
  summarise(
    Total_Rows = n(),
    Missing_HF = sum(is.na(HF_value)),
    Percent_Missing = round((Missing_HF / Total_Rows) * 100, 2)
  )
print(activity_session_day_na_summary)


person_na_summary <- data %>%
  group_by(person) %>%
  summarise(
    Total_Rows = n(),
    Missing_HF = sum(is.na(HF_value)),
    Percent_Missing = round((Missing_HF / Total_Rows) * 100, 2)
  )
print(person_na_summary)

activity_session_day_sensor_na_summary <- data %>%
  group_by(activity, session, day, HF_type) %>%
  summarise(
    Total_Rows = n(),
    Missing_HF = sum(is.na(HF_value)),
    Percent_Missing = round((Missing_HF / Total_Rows) * 100, 2)
  )
print(activity_session_day_sensor_na_summary, n =120)
```


```{r data cleaning}
#clean dataset
#drop NAs, order dataset by person, sensor, day, session, activity
data_clean <- data %>%
  drop_na() %>%
  mutate(
    person = as.factor(person),
    day = factor(day, levels = c("1", "2")), 
    session = factor(session, levels = c("1", "2", "3", "4")),
    activity = factor(activity, levels = c("sitting", "standing", "walking")),
    sex = as.factor(sex),
    HF_type = as.factor(HF_type),
    sensor_num = as.numeric(HF_type),
    session_num = as.numeric(session),
    Block_ID = as.factor(paste(person, day, session, activity, sep="_"))
  ) %>%
  arrange(person, HF_type, day, session, activity)
```

```{r output baseline for dummy encoding}
# baselines
levels(data_clean$sex)[1] #F
levels(data_clean$activity)[1] #sitting
levels(data_clean$day)[1] #1
levels(data_clean$session)[1] #1
levels(data_clean$HF_type)[1]  #HF_16
```


```{r Check Linearity Visually}
# make session to be continuous variable for now
# spaghetti plot + faceting
data_sitting <- data_clean %>%
  filter(activity == "sitting")

ggplot(data_sitting, aes(x = session_num, y = HF_value, color = sex)) +
  geom_point(alpha = 0.1) +
  
  # Flexible LOESS curves (Dashed lines, one for M and one for F)
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, linetype = "dashed", size = 0.5) +
  
  # Straight Linear regression lines (Solid lines, one for M and one for F)
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linetype = "solid", size = 0.5) +
  
  # Grid: Rows = Activity, Columns = Sensor
  facet_grid(day ~ HF_type) +
  
  # Manually set colors for clarity (e.g., Males = Blue, Females = Red)
  scale_color_manual(values = c("M" = "blue", "F" = "red")) +
  
  theme_minimal() +
  labs(
    title = "(Sitting) Heat Flow Over Time by Day, Session, Sensor and Sex",
    subtitle = "Solid lines = Linear Fit (lm) | Dashed lines = Flexible Fit (loess)",
    x = "Session (1 to 4)",
    y = "Heat Flow (HF_value)",
    color = "Sex"
  )
```


```{r plot to justify fixed effect}

data_day1 <- data_clean %>% filter(day == "1")

ggplot(data_day1, aes(x = session_num, y = HF_value, color = sex)) +
  geom_point(alpha = 0.1) + 
  
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, linetype = "dashed", size = 0.5) +
  
  # Straight Linear regression lines (Solid lines, one for M and one for F)
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linetype = "solid", size = 0.5) +
  
  facet_grid(activity ~ HF_type) +
  
  scale_color_manual(values = c("M" = "blue", "F" = "red")) +
  
  theme_minimal() +
  labs(
    title = "(Day 1)Heat Flow Over Time by Activity, Sensor and Sex",
    subtitle = "Solid lines = Linear Fit | Dashed = LOESS | Grid = Activity+Day vs Sensor",
    x = "Session (1 to 4)",
    y = "Heat Flow (HF_value)",
    color = "Sex"
  ) 

#Since all variables are categorical, we don't need to check for linearity assumption.
#But above plot indicates there is fixed effect from every categorical variables. We should include all of them in our model.
#(no need to test if each fixed effect exists, just from plot and intuition, also since fixed effects are the same, in later models we can use REML which doesn't influence comparison)

```



## Section 5: Results

## Section 5.1: Initial Model

```{r naive linear model}

naive_lm <- lm(HF_value ~ session + activity + day + HF_type + sex, data = data_clean)

summary(naive_lm)
#Also based on the significance for each coefficients, we should include all of the predictors.

```

```{r check assumptions naive linear model}
par(mfrow = c(2, 2))
plot(naive_lm)
```


```{r residuals distribution before log-transform}
hist(resid(naive_lm), breaks = 30, probability = TRUE, col = "lightblue", main = "Histogram of Residuals", xlab = "Residual Error")
lines(density(resid(naive_lm)), col = "red", lwd = 2)
```

```{r log transformed naive linear model}
log_naive_lm <- lm(log(HF_value) ~ session + activity + day + HF_type + sex, data = data_clean)
summary(log_naive_lm)
par(mfrow = c(2, 2))
plot(log_naive_lm)
```

```{r check assumptions, include = TRUE, fig.width=10, fig.height=10, fig.align='center',out.width="100%"}


par(mfrow = c(2, 2), mar = c(4, 4, 3, 2), oma = c(5, 0, 3, 0))
#2 rows 2 columns
#oma = outer margins to have space for caption

#NAIVE (row 1)
# Plot 2: Normal Q-Q
plot(naive_lm, which = 2, main = "Model 0 (Raw): Normality (QQ)", sub = "", caption = "") #sub means do not show normal subtitles, manually create my own
# Plot 3: Scale-Location (Spread-Location)
plot(naive_lm, which = 3, main = "Model 0 (Raw): Scale-Location", sub = "", caption = "") #caption means do not show normal subtitles - replace with main

#LOG (row 2)
# Plot 2: Normal Q-Q
plot(log_naive_lm, which = 2, main = "Model 1 (Log): Normality (QQ)", sub = "", caption = "")
# Plot 3: Scale-Location (Spread-Location)
plot(log_naive_lm, which = 3, main = "Model 1 (Log): Scale-Location", sub = "", caption = "")


caption_text<- "Figure 1: The QQ-plot of the Model 0 (the raw (untransformed) model) shows an approximately normal model with slightly heavy tails, but the Scale-Location plot shows heteroscedastic residuals, with variance increasing as fitted values increase. Model 1 (the log-transformed model) corrects the right heavy tail but the left tail deviates more, and the Scale-Location plot shows fairly homoscedastic residuals."

wrapped_caption <- paste(strwrap(caption_text, width = 150), collapse = "\n")

mtext("Figure 1: Diagnostic Comparison of Raw & Log Linear Models", 
      side = 3, line = 0, outer = TRUE, cex = 1.2, font = 2)

mtext(wrapped_caption, 
      side = 1, line = 2, outer = TRUE, cex = 0.8)
```

As seen in Figures 1 and 2, applying a simple linear model (Model 0: $\text{HF\_value} \sim \text{session + activity + day + HF\_type + sex}$) shows that the residuals are approximately normal (with heavy tails), but heteroscedastic. Transforming the response variable (Model 1: $\log(\text{HF\_value})$) resulted in homoscedasticity and corrected the heavy right tail, but introduced more left-skew in the residuals. We prioritize homoscedasticity and use Model 1 (the log model) going forward, to ensure reliable parameter estimation and significance testing.


```{r check assumptions math, include = TRUE, fig.width=10, fig.height=10, fig.align='center',out.width="100%"}
fit_naive <- fitted(naive_lm)
res_naive <- residuals(naive_lm)

fit_log <- fitted(log_naive_lm)
res_log <- residuals(log_naive_lm)

# Calculate Correlation for Heteroscedasticity - closer to 0 means heteroscedastic
het_corr_naive <- cor(fit_naive, abs(res_naive))
het_corr_log <- cor(fit_log, abs(res_log))

#Calculate Correlation for Normality (Q-Q Correlation) - closer to 1 means more normal
qq_naive <- cor(sort(res_naive), qnorm(ppoints(length(res_naive))))
qq_log <- cor(sort(res_log), qnorm(ppoints(length(res_log))))

log_naive_comparison <- data.frame(
  Metric = c("Normality", "Heteroscedasticity"),
  Naive_Model = c(round(qq_naive, 4), round(het_corr_naive, 4)),
  Log_Model   = c(round(qq_log, 4), round(het_corr_log, 4))
)
caption_table <- "Table 1: Mathematical diagnostics comparison between Model 0 (raw) and Model 1 (log-transformed model), comparing normality (correlation between actual and theoretically normal residuals, with values closer to 1 indicating normality) and heteroscedasticity (correlation between fitted values and residuals, with values closer to 0 indicating heteroscedasticity). While Model 1 shows slightly worse normality, it is significantly closer to homoscedasticity."


kable(log_naive_comparison, 
      col.names = c("Metric", "Model 0 (Raw)", "Model 1 (Log)"),
      align = "lrr",
      booktabs = TRUE) %>% #, #right aligns numbers
    #   caption = caption_table) %>%
    kable_styling(#latex_options = "hold_position", 
                full_width = FALSE) %>%
    add_header_above(c("Table 1: Mathematical Diagnostic Comparison of Model 0 & Model 1" = 3), 
                   bold = TRUE, align = "l") %>%
    row_spec(0, bold = TRUE) %>%
    footnote(general = paste0("\\\\footnotesize{", caption_table, "}"),
           general_title = "", 
           threeparttable = TRUE,
           escape = FALSE)

```


## Section 5.2: Fixed Effects

```{r check what fixed effects to include}
mdl1_log <- gls(
  log(HF_value) ~ day + session + activity + HF_type + sex,
  data   = data_clean,
  method = "ML"
)
mdl1_log_no_day <- gls(
  log(HF_value) ~ session + (activity) +  (HF_type) +  (sex), 
  data   = data_clean,
  method = "ML"
)
mdl1_log_no_session <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex), 
  data   = data_clean,
  method = "ML"
)
mdl1_log_no_activity <- gls(
  log(HF_value) ~  (day) +  (session) +  (HF_type) +  (sex), 
  data   = data_clean,
  method = "ML"
)
mdl1_log_no_sensor <- gls(
  log(HF_value) ~  (day) +  (session) +  (activity) +  (sex), 
  data   = data_clean,
  method = "ML"
)
mdl1_log_no_sex <- gls(
  log(HF_value) ~  (day) +  (session) +  (activity) +  (HF_type), 
  data   = data_clean,
  method = "ML"
)

#helper function to get anova outputs
get_baseline_metrics <- function(orig, name) {
  data.frame(
    Model = name,
    Df = attr(logLik(orig), "df"),
    AIC = round(AIC(orig)),
    BIC = round(BIC(orig)),
    LogLik = round(as.numeric(logLik(orig))),
    stringsAsFactors = FALSE
  )
}

#note that type=marginal means it is an F test!!
get_anova_outputs <- function(orig, new, name) {
  res <- anova(orig, new) #, type = "marginal" doesnt matter if pass 2 models; uses chi-squared
  data.frame(
    Model = name,
    Df = attr(logLik(new), "df"),
    AIC = round(res$AIC[2]),
    BIC = round(res$BIC[2]),
    LogLik = round(res$logLik[2]),
    LRT_Stat = round(res$L.Ratio[2],2),
    p_value = res$`p-value`[2]
  )
}

anova_summary_manual_mdl1 <- bind_rows(
  get_baseline_metrics(mdl1_log, "Model 1 (Full Model)"),
  get_anova_outputs(mdl1_log, mdl1_log_no_day, "Drop Day"),
  get_anova_outputs(mdl1_log, mdl1_log_no_session, "Drop Session"),
  get_anova_outputs(mdl1_log, mdl1_log_no_activity, "Drop Activity"),
  get_anova_outputs(mdl1_log, mdl1_log_no_sensor, "Drop Sensor"),
  get_anova_outputs(mdl1_log, mdl1_log_no_sex, "Drop Sex")
)

print(anova_summary_manual_mdl1)

anova_summary_manual_mdl1_cleaned <- anova_summary_manual_mdl1 %>%
  mutate(
    LRT_Stat = ifelse(is.na(LRT_Stat), "--", as.character(LRT_Stat)),
    p_value  = ifelse(is.na(p_value), "--", 
                      ifelse(p_value < 0.0001, "< 0.0001", as.character(p_value)))
  )

caption_table <- "Table 2: Model selection of fixed effects via backwards elimination. Model 1 (the 'Full Model') includes all predictors. Each subsequent row represents a model where only one term was dropped from Model 1. The LRT Statistic and P-Value are derived from Likelihood Ratio Tests comparing the reduced model to Model 1. Note all models were fit using Maximum Likelihood (ML), to compare across models."


col_labels <- c("Model Version", 
                "\\shortstack{Degrees of\\\\Freedom}", 
                "AIC", 
                "BIC", 
                "Log-Likelihood", 
                "\\shortstack{Likelihood Ratio\\\\Test Stat}", 
                "P-Value")

```

```{r run table 2, include = TRUE}
kable(anova_summary_manual_mdl1_cleaned, 
      format="latex",
      col.names = col_labels,
      align = "lrrrrrr",
      booktabs = TRUE,#, #right aligns numbers
      escape = FALSE) %>% 
    #   caption = caption_table) %>%
    kable_styling(latex_options = "scale_down", 
                full_width = FALSE) %>%
    add_header_above(c("Table 2: Backwards Selection of Fixed Effects: Likelihood Ratio Test Summary" = 7), 
                   bold = TRUE, align = "l") %>%
    row_spec(0, bold = TRUE) %>% #highlight main row
    row_spec(1, italic = TRUE) %>%
    footnote(
           # This creates a box exactly the width of a standard page (15cm)
           # The caption will wrap inside this box and NOT shrink the table.
           general = paste0("\\\\begin{minipage}{20cm}\\\\vspace{2mm}\\\\footnotesize{", caption_table, "}\\\\end{minipage}"),
           general_title = "", 
           threeparttable = FALSE, 
           escape = FALSE)
```

As seen in Table 2, although Model 1 includes all fixed effects ($\log(\text{HF\_value}) \sim \text{session + activity + day + sensor + sex}$), a backwards selection process was conducted to ensure parsimony. We evaluated each variable by comparing the full model to nested models where one term was removed, using the Likelihood Ratio Test (LRT) and changes in information criteria (AIC and BIC).

At first glance, all variables are statistically significant except for the model dropping the session variable ($\log(\text{HF\_value}) \sim \text{ activity + day + sensor + sex}$), as the p-value from the likelihood ratio test of this model versus the full model is not below 0.05 (p=0.1691). Additionally, although the models that drop activity and drop sex are technically statistically significant (p-values < 0.01), these models show a unique characteristic of having slightly higher AICs than the full model, but lower or similar BICs. This discrepancy implies that although these variables improve prediction, the improvement may not be large enough to justify including the variables. In this case (and similar cases further in the analysis), the variables are included but will be re-examined in final model pruning.

As such, the final fixed effect model was Model 2 ($\log(\text{HF\_value}) \sim \text{ activity + day + sensor + sex}$).

## Section 5.3: Interaction Terms

We further examined fixed effects by examining the value of adding all possible interaction terms to the model, as shown in Table 3. Testing adding each interaction term against against Model 2 ($\text{HF\_value} \sim \text{ activity + day + sensor + sex}$), we saw that adding the following interaction terms results in models with worse (or similar) AIC & BIC than the current model without interaction terms, and thus did not add them to the model: Activity & Day, Activity & Sex, Session & Sex, Day & Sex, Day & Session. 

Technically, adding the interaction of Day & Session results in a p-value of 0.0118, below 0.05. However, this interaction increased the BIC and barely decreases the AIC, so we chose not to include it in the model. Note that the interaction of Acitivity & Session was not tested - because there is no data for certain interactions (ex. walking & session 2, walking & session 4), the model cannot run.

The interaction of Sensor with the other possible variables (Day, Activity, Sex, and Session) all have low p-values (<0.001) and result in decreased AIC, although two interaction terms (Sensor and Sex, and Sensor and Session) result in increased BIC. Table 2 showed a similar story - both Session and Sex decreased AIC slightly, but increased BIC. Similar to Section 5.1, we include these interaction terms for now and will prune the model at the end.

```{r interactive tests}

#sex*session
mdl2_log_no_session_interact_sex_session<- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (sex * session) ,
  data   = data_clean,
  method = "ML"
)

#day*sex
dl2_log_no_session_interact_sex_day<- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (sex * day) ,
  data   = data_clean,
  method = "ML"
)

#day*activity
dl2_log_no_session_interact_day_activity<- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (activity * day) ,
  data   = data_clean,
  method = "ML"
)


#HF_type*day
mdl2_log_no_session_interact_sensor_day <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) ,
  data   = data_clean,
  method = "ML"
)

#HF_type*activity
mdl2_log_no_session_interact_sensor_activity <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * activity) ,
  data   = data_clean,
  method = "ML"
)

#HF_type*sex
mdl2_log_no_session_interact_sensor_sex <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * sex) ,
  data   = data_clean,
  method = "ML"
)


#HF_type*session
mdl2_log_no_session_interact_sensor_session <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * session) ,
  data   = data_clean,
  method = "ML"
)


#activity*sex
mdl2_log_no_session_interact_activity_sex <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (activity * sex) ,
  data   = data_clean,
  method = "ML"
)

#sex*session
mdl2_log_no_session_interact_session_sex <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (session * sex) ,
  data   = data_clean,
  method = "ML"
)

#day*session
mdl2_log_no_session_interact_session_day <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (session * day) ,
  data   = data_clean,
  method = "ML"
)

#session*activity
#error: Error in `glsEstimate()`:! computed "gls" fit is singular, rank 18
#we are missing some data (NAs) - ex walking for session 2 & 4, likely the reason?
# mdl2_log_no_session_interact_session_activity <- gls(
#   log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (session * activity) ,
#   data   = data_clean,
#   method = "ML"
# )

anova_summary_manual_mdl1_nosession_interact1 <- bind_rows(
  get_baseline_metrics(mdl1_log_no_session, "Model 2 (No Interaction Terms)"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_day, "Interact Sensor and Day"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_activity, "Interact Sensor and Activity"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_sex, "Interact Sensor and Sex"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_session, "Interact Sensor and Session"),
  get_anova_outputs(mdl1_log_no_session, dl2_log_no_session_interact_day_activity, "Interact Activity and Day"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_activity_sex, "Interact Activity and Sex"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_sex_session, "Interact Session and Sex"),
  get_anova_outputs(mdl1_log_no_session, dl2_log_no_session_interact_sex_day, "Interact Day and Sex"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_session_day, "Interact Day and Session")
)


print(anova_summary_manual_mdl1_nosession_interact1)


anova_summary_manual_mdl1_nosession_interact1_cleaned <- anova_summary_manual_mdl1_nosession_interact1 %>%
  mutate(
    LRT_Stat = ifelse(is.na(LRT_Stat), "--", as.character(LRT_Stat)),
    p_value  = ifelse(is.na(p_value), "--", 
                      ifelse(p_value < 0.0001, "< 0.0001", as.character(p_value)))
  )

print("hi")
print(ncol(anova_summary_manual_mdl1_nosession_interact1_cleaned))

caption_table <- "Table 3: The top row (Model 2) is the row without any interaction terms. Each subsequent row represents a model where only one interaction term was added to Model 2. The LRT Statistic and P-Value are derived from Likelihood Ratio Tests comparing the new model to Model 2 via ANOVA. Note all models were fit using Maximum Likelihood (ML), to compare across models."

# Clean the Data Frame

col_labels <- c("Model Version", 
                "\\shortstack{Degrees of\\\\Freedom}", 
                "AIC", 
                "BIC", 
                "Log-Likelihood", 
                "\\shortstack{Likelihood Ratio\\\\Test Stat}", 
                "P-Value")

```


```{r run table 3, include = TRUE}

kable(anova_summary_manual_mdl1_nosession_interact1_cleaned, 
      format="latex",
      col.names = col_labels,
      align = "lrrrrrr",
      booktabs = TRUE,#, #right aligns numbers
      row.names = FALSE,
      escape = FALSE) %>% 
    #   caption = caption_table) %>%
    kable_styling(latex_options = "scale_down", 
                full_width = FALSE) %>%
    add_header_above(c("Table 3: Interaction Terms: Likelihood Ratio Test Summary" = 7), 
                   bold = TRUE, align = "l") %>%
    row_spec(0, bold = TRUE) %>% #highlight main row
    row_spec(1, italic = TRUE) %>%
    footnote(
           # This creates a box exactly the width of a standard page (15cm)
           # The caption will wrap inside this box and NOT shrink the table.
           general = paste0("\\\\begin{minipage}{\\\\linewidth}\\\\vspace{2mm}\\\\footnotesize{", caption_table, "}\\\\end{minipage}"),
           general_title = "", 
           threeparttable = TRUE, 
           escape = FALSE)
```

This results in Model 3: ($\log(\text{HF\_value}) \sim \text{activity + day + sensor + sex + sensor * activity + sensor * day + sensor * sex + sensor * session + session}$). Note that session is added back into the model automatically, as the interaction of sensor and session is included, due to the principle of marginality. Table 4 shows the results of checking if removing any of the four interaction terms improves Model 3. This showed a similar story to the individual tests of adding interaction terms - the inclusion of the interactions of Sensor and Sex, and Sensor and Session, slightly decrease AIC but increase BIC. 

```{r}
#HF_type*day, HF_type * activity, HF_type * sex, HF_type * session, day * session
mdl3 <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (session) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type * session),
  data   = data_clean,
  method = "ML"
)

#remove hf_type * day
mdl3a <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) +(session) +(HF_type * activity) + (HF_type * sex) + (HF_type * session),
  data   = data_clean,
  method = "ML"
)

#remove hf_type * activity
mdl3b <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) +(session) + (HF_type * day) + (HF_type * sex) + (HF_type * session),
  data   = data_clean,
  method = "ML"
)

#remove hf_type * sex
mdl3c <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) +(session) + (HF_type * day) + (HF_type * activity) + (HF_type * session),
  data   = data_clean,
  method = "ML"
)

#remove hf_type * session, and session
mdl3d <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) ,
  data   = data_clean,
  method = "ML"
)

anova_summary_manual_mdl1_nosession_interact2 <- bind_rows(
  get_baseline_metrics(mdl3, "Model 3 (Four Interaction Terms)"),
  get_anova_outputs(mdl3, mdl3a, "Drop Sensor and Day"),
  get_anova_outputs(mdl3, mdl3b, "Drop Sensor and Activity"),
  get_anova_outputs(mdl3, mdl3c, "Drop Sensor and Sex"),
  get_anova_outputs(mdl3, mdl3d, "Drop Sensor and Session")
)
print(anova_summary_manual_mdl1_nosession_interact2)


caption_table <- "Table 4: Model selection of interaction terms via backwards elimination. Model 3 includes the four interaction terms previously noted. Each subsequent row represents a model where only one interaction term was dropped from Model 3. Note that for the last row, Drop Sensor and Session, the fixed effect Session is dropped as well (Session was only included in Model 3 because the interaction of Sensor and Session was added). The LRT Statistic and P-Value are derived from Likelihood Ratio Tests comparing the reduced model to Model 3. Note all models were fit using Maximum Likelihood (ML), to compare across models."


col_labels <- c("Model Version", 
                "\\shortstack{Degrees of\\\\Freedom}", 
                "AIC", 
                "BIC", 
                "Log-Likelihood", 
                "\\shortstack{Likelihood Ratio\\\\Test Stat}", 
                "P-Value")

```

```{r run table 4, include = TRUE}

kable(anova_summary_manual_mdl1_nosession_interact2, 
      format="latex",
      col.names = col_labels,
      align = "lrrrrrr",
      booktabs = TRUE,#, #right aligns numbers
      escape = FALSE) %>% 
    #   caption = caption_table) %>%
    kable_styling(latex_options = "scale_down", 
                full_width = FALSE) %>%
    add_header_above(c("Table 4: Backwards Selection of Interaction Terms: Likelihood Ratio Test Summary" = 7), 
                   bold = TRUE, align = "l") %>%
    row_spec(0, bold = TRUE) %>% #highlight main row
    row_spec(1, italic = TRUE) %>%
    footnote(
           # This creates a box exactly the width of a standard page (15cm)
           # The caption will wrap inside this box and NOT shrink the table.
           general = paste0("\\\\begin{minipage}{15cm}\\\\vspace{2mm}\\\\footnotesize{", caption_table, "}\\\\end{minipage}"),
           general_title = "", 
           threeparttable = TRUE, 
           escape = FALSE)
```

## Section 5.4: Random Intercepts

## TO DO: ADD JUSTIFICIATION FOR LMM

```{r Boxplot of Residuals by Person, include = TRUE}

# residuals added to dataframe
data_clean$residuals_model1 <- resid(log_naive_lm)

ggplot(data_clean, aes(x = person, y = residuals_model1, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  # red line: perfect independent residuals should be centered at 0
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Residuals by Person for Model 1, fixed-effect-only model",
    x = "Participant",
    y = "Residual Error"
  )


data_clean$residuals_model3 <- resid(mdl3)

ggplot(data_clean, aes(x = person, y = residuals_model3, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  # red line: perfect independent residuals should be centered at 0
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Motivation for LMM: Residuals by Person (Model 3)",
    subtitle = "Systematic shifts away from zero prove that errors within an individual are correlated.",
    x = "Participant",
    y = "Residual Error (Observed - Predicted)"
  )
```



To determine if there is inter-individual variability, we added a random intercept for each person, creating Model 4: ($\log(\text{HF\_value}) \sim \text{activity + day + sensor + sex + sensor * activity + sensor * day + sensor * sex + sensor * session + session}$). This assumes that all individuals have unique baselines for heat flow, regardless of if they are performing the same activity and/or are at the same temperature. Note the models in this section (5.4) were run using REML to ensure estimates of variance components are not biased, unless explicitly stated otherwise.

```{r random intercept model}

mdl3_reml <- gls(log(HF_value) ~ (day + session + activity + sex) * HF_type, 
          data = data_clean, 
          method = "REML")

mdl4a_reml <- lme(fixed = log(HF_value) ~ (day + session + activity + sex) * HF_type, 
          random = ~ 1 | person, #unique intercept for each person
          data = data_clean, 
          method = "REML")

summary(mdl4a_reml)

anova(mdl3_reml, mdl4a_reml)
```

```{r random intercept model person}
print(VarCorr(mdl4a_reml))

mdl4a_var <- as.numeric(VarCorr(mdl4a_reml))
print(mdl4a_var )

#these are the variances
print(mdl4a_var[1])
print(mdl4a_var[2])

mdl4a_var_person <- mdl4a_var[1]
mdl4a_var_residual <- mdl4a_var[2]

mdl4a_icc <- mdl4a_var_person/(mdl4a_var_person+mdl4a_var_residual)
mdl4a_icc_res <- mdl4a_var_residual/(mdl4a_var_person+mdl4a_var_residual)
print(mdl4a_icc)
print(mdl4a_icc_res)
```

The Intraclass Correlation Coefficient (ICC) for Model 4 was 0.0791, implying that ~8% of the variance is due to differences between people (and 92% is unexplained). This suggests there is some inter-individual variability. Table 5 shows that this also reduces AIC and BIC significantly, 

```{r}
anova_mdl3_mdl4 <- bind_rows(
  get_baseline_metrics(mdl3_reml, "Model 3"),
  get_anova_outputs(mdl3_reml, mdl4a_reml, "Model 4")
)


caption_table <- "Table 5: Impact of adding a random intercept for person (Model 4) to the fixed-effect only model (Model 3). Note this uses REML to compare models, as fixed effects are not changing. As such, the final p-values should be divided by 2. Given the p-value is already extremely low, this has no impact on the takeaway that we should include the Random Intercept for Person. Note models were fit using Restricted Maximum Likelihood (REML), as fixed effects are the same between models."


col_labels <- c("Model Version", 
                "\\shortstack{Degrees of\\\\Freedom}", 
                "AIC", 
                "BIC", 
                "Log-Likelihood", 
                "\\shortstack{Likelihood Ratio\\\\Test Stat}", 
                "P-Value")

```

```{r create table 5, include = TRUE}

kable(anova_mdl3_mdl4, 
      format="latex",
      col.names = col_labels,
      align = "lrrrrrr",
      booktabs = TRUE,#, #right aligns numbers
      escape = FALSE) %>% 
    #   caption = caption_table) %>%
    kable_styling(latex_options = "scale_down", 
                full_width = FALSE) %>%
    add_header_above(c("Table 5: Likelihood Ratio Test for Random Intercept for Person" = 7), 
                   bold = TRUE, align = "l") %>%
    row_spec(0, bold = TRUE) %>% #highlight main row
    footnote(
           # This creates a box exactly the width of a standard page (15cm)
           # The caption will wrap inside this box and NOT shrink the table.
           general = paste0("\\\\begin{minipage}{15cm}\\\\vspace{2mm}\\\\footnotesize{", caption_table, "}\\\\end{minipage}"),
           general_title = "", 
           threeparttable = TRUE, 
           escape = FALSE)
```

```{r random intercept model person activity}

mdl4b_reml <- lme(fixed = log(HF_value) ~ (day + session + activity + sex) * HF_type, 
          random = ~ 1 | person/activity, #unique intercept for each person
          data = data_clean, 
          method = "REML")

print(VarCorr(mdl4b_reml))

mdl4b_var <- as.numeric(VarCorr(mdl4b_reml))
print(mdl4b_var )

#these are the variances
print(mdl4b_var[1])
print(mdl4b_var[2])
print(mdl4b_var[3])
print(mdl4b_var[4])
print(mdl4b_var[5])
print(mdl4b_var[6])


mdl4b_var_person <- mdl4b_var[2]
mdl4b_var_activity <- mdl4b_var[4]
mdl4b_var_residual <- mdl4b_var[5]
mdl4b_var_total <- mdl4b_var_person + mdl4b_var_activity + mdl4b_var_residual

mdl4b_icc_person <- round(mdl4b_var_person / mdl4b_var_total, 4)
mdl4b_icc_activity <- round(mdl4b_var_activity / mdl4b_var_total, 4)
mdl4b_icc_res <- round(mdl4b_var_residual / mdl4b_var_total, 4)

print("icc person")
print(mdl4b_icc_person)

print("icc activity")
print(mdl4b_icc_activity)
print("icc residual")
print(mdl4b_icc_res)
```

We then examined if a nested activity-level random effect was necessary. The ICC for person remains the same as in model 4 (0.0791), and the ICC of activity (within person) is 0, implying adding activity as a random intercept adds no value - the effect of activity does not vary randomly between people.

```{r random intercept model person sensor}

mdl4c_reml <- lme(fixed = log(HF_value) ~ (day + session + activity + sex) * HF_type, 
          random = ~ 1 | person/HF_type, #unique intercept for each person
          data = data_clean, 
          method = "REML")

print(VarCorr(mdl4c_reml))

mdl4c_var <- as.numeric(VarCorr(mdl4c_reml))
print(mdl4c_var )

#these are the variances
print(mdl4c_var[1])
print(mdl4c_var[2])
print(mdl4c_var[3])
print(mdl4c_var[4])
print(mdl4c_var[5])
print(mdl4c_var[6])


mdl4c_var_person <- mdl4c_var[2]
mdl4c_var_sensor <- mdl4c_var[4]
mdl4c_var_residual <- mdl4c_var[5]
mdl4c_var_total <- mdl4c_var_person + mdl4c_var_sensor + mdl4c_var_residual

mdl4c_icc_person <- round(mdl4c_var_person / mdl4c_var_total, 4)
mdl4c_icc_sensor <- round(mdl4c_var_sensor / mdl4c_var_total, 4)
mdl4c_icc_res <- round(mdl4c_var_residual / mdl4c_var_total, 4)

print("icc person")
print(mdl4c_icc_person)

print("icc sensor")
print(mdl4c_icc_sensor)

print("icc residual")
print(mdl4c_icc_res)
```

```{r random intercept model person sensor but remove person!}

mdl4d_reml <- lme(fixed = log(HF_value) ~ (day + session + activity + sex) * HF_type, 
          random = ~ 1 | HF_type, #unique intercept for each person
          data = data_clean, 
          method = "REML")

print(VarCorr(mdl4d_reml))

mdl4d_var <- as.numeric(VarCorr(mdl4d_reml))
print(mdl4d_var )

#these are the variances
print(mdl4d_var[1])
print(mdl4d_var[2])
print(mdl4d_var[3])
print(mdl4d_var[4])


mdl4d_var_sensor <- mdl4d_var[1]
mdl4d_var_residual <- mdl4d_var[2]
mdl4d_var_total <- mdl4d_var_sensor + mdl4d_var_residual 

mdl4d_icc_sensor <- round(mdl4d_var_sensor / mdl4d_var_total, 4)
mdl4d_icc_res <- round(mdl4d_var_residual / mdl4d_var_total, 4)

print("icc sensor")
print(mdl4d_icc_sensor)
print("icc residaul")
print(mdl4d_icc_res)
```

```{r table 5 anova check}
print(anova(mdl3_reml, mdl4a_reml))
print(anova(mdl4a_reml, mdl4c_reml))
print(anova(mdl4c_reml, mdl4d_reml))
```


We then examined if a nested sensor-level random intercept was necessary (Model 5). The ICC for person is now 0.0184, lower than in model 4 (0.0791), and the ICC of sensor (within person) is 0.3772, with 60% of the variance unexplained. This implies that the major source of clustering is at the sensor level, and not individual level.

Removing the person random intercept (resulting in a sensor-only random intercept model) results in the significantly lower ICC for sensor (0.0159), with 98% of the variance left unexplained. This implies that the sensor by itself is not the primary driver of variance; instead, there is a shared baseline amongst sensors belonging to the same individual. Thus, we keep the nested sensor-level random intercept model (model 5).

```{r anova sensor ri}

anova_mdl5 <- anova(mdl4c_reml, mdl4d_reml)
anova_mdl5_clean <- data.frame(
  Model = c("Model 5: Random Intercept: Person and Nested Sensor", "Random Intercept: Sensor only"),
  Df = as.numeric(anova_mdl5$df),
  AIC = round(as.numeric(anova_mdl5$AIC)),
  BIC = round(as.numeric(anova_mdl5$BIC)),
  LogLik = round(as.numeric(anova_mdl5$logLik)),
  LRT_Stat = c("--", round(anova_mdl5$L.Ratio[2], 2)),
  p_value = anova_mdl5$`p-value`[2])

```

```{r create table6, include = FALSE}

# caption_table <- "Table 6: Model 5 has a random intercept for person, and a nested random intercept for sensor. Removing the random intercept for person from Model 5 (resulting in a model that only has a random intercept for sensor) significantly decreases the AIC and BIC, even though person does not have a large ICC in model 5. Note models were fit using Restricted Maximum Likelihood (REML), as fixed effects are the same between models."


# col_labels <- c("Model Version", 
#                 "\\shortstack{Degrees of\\\\Freedom}", 
#                 "AIC", 
#                 "BIC", 
#                 "Log-Likelihood", 
#                 "\\shortstack{Likelihood Ratio\\\\Test Stat}", 
#                 "P-Value")

# kable(anova_mdl5_clean, 
#       format="latex",
#       col.names = col_labels,
#       align = "lrrrrrr",
#       booktabs = TRUE,#, #right aligns numbers
#       escape = FALSE) %>% 
#     #   caption = caption_table) %>%
#     kable_styling(latex_options = "scale_down", 
#                 full_width = FALSE) %>%
#     add_header_above(c("Table 5: Likelihood Ratio Test for Random Intercept for Person" = 7), 
#                    bold = TRUE, align = "l") %>%
#     row_spec(0, bold = TRUE) %>% #highlight main row
#     footnote(
#            # The caption will wrap inside this box and NOT shrink the table.
#            general = paste0("\\\\begin{minipage}{20cm}\\\\vspace{2mm}\\\\footnotesize{", caption_table, "}\\\\end{minipage}"),
#            general_title = "", 
#            threeparttable = TRUE, 
#            escape = FALSE)
```


## Section 5.5: Random Slopes

The current model has random intercepts for person and sensor (nested), implying that measurements from the same sensor are correlated, and sensors belonging to the same person are correlated. However, it assumes that the impact of temperature (the day variable) is identical for all individuals. We tested day as a random slope (given person) to determine if the effect of day varies by person, meaning some people adapt to the temperature environment differently than others.

Adding day as a random slope (Model 6) results in 45% unexplained variance, compared to 60% from Model 5. The ICC for day is 0.186, which is fairly large; ~19% of the variation in heat flow measurements is explained by the fact that people have different heat flow reactions depending on the external temperature. Additionally, the correlation between the person-level intercept and the day-level random slope is -0.884. This implies that the higher a person's heat flow was on day 1 (hot day), the steeper their drop was on day 2 (cold day), suggesting convergence.

The ICC for person also increased to 0.0419 (compared to 0.0184 from model 5), suggesting that the noise from people reacting differently to temperature masked their baseline differences. The ICC for sensor decreased to 0.3241 (from 0.3772 in model 5), suggesting that temperature explained some of the heat flow variance that was previously attributed to sensor.

```{r add random slope}
#mdl 4c is model 5

mdl6_reml <- lme(
  fixed = log(HF_value) ~ (day + session + activity + sex) * HF_type,
  random = list(
    person = ~ 1 + day,
    HF_type = ~ 1
  ), 
  data = data_clean, 
  method = "REML"
)

print(anova(mdl4c_reml, mdl6_reml))

print(VarCorr(mdl6_reml))

mdl6_var <- as.numeric(VarCorr(mdl6_reml))
print(mdl6_var)

#these are the variances
print(mdl6_var[1])
print(mdl6_var[2])
print(mdl6_var[3])
print(mdl6_var[4])
print(mdl6_var[5])
print(mdl6_var[6])
print(mdl6_var[15]) #correlation


mdl6_var_person <- mdl6_var[2]
mdl6_var_day <- mdl6_var[3]
mdl6_var_sensor <- mdl6_var[5]
mdl6_var_residual <- mdl6_var[6]
mdl6_var_total <- mdl6_var_person + mdl6_var_sensor + mdl6_var_day+ mdl6_var_residual

mdl6_icc_person <- round(mdl6_var_person / mdl6_var_total, 4)
mdl6_icc_sensor <- round(mdl6_var_sensor / mdl6_var_total, 4)
mdl6_icc_day <- round(mdl6_var_day / mdl6_var_total, 4)
mdl6_icc_res <- round(mdl6_var_residual / mdl6_var_total, 4)

mdl6_corr_day_person <- mdl6_var[15]

print("icc person mdl6")
print(mdl6_icc_person)

print("icc day mdl6")
print(mdl6_icc_day)

print("icc sensor mdl6")
print(mdl6_icc_sensor)

print("icc residual mdl6")
print(mdl6_icc_res)

print("icc person mdl4c")
print(mdl4c_icc_person)

print("icc sensor mdl4c")
print(mdl4c_icc_sensor)

print("icc residual mdl4c")
print(mdl4c_icc_res)


print("corr day person")
print(mdl6_corr_day_person)
```

```{r anova mdl6 rs}

anova_mdl6 <- anova(mdl4c_reml, mdl6_reml)
anova_mdl6_clean <- data.frame(
  Model = c("Model 5: Random Intercept: Person and Nested Sensor", "Model 6: Add Random Slope for Day (Given Person)"),
  Df = as.numeric(anova_mdl6$df),
  AIC = round(as.numeric(anova_mdl6$AIC)),
  BIC = round(as.numeric(anova_mdl6$BIC)),
  LogLik = round(as.numeric(anova_mdl6$logLik)),
  LRT_Stat = c("--", round(anova_mdl6$L.Ratio[2], 2)),
  p_value = anova_mdl6$`p-value`[2])

print(anova_mdl6)
```

```{r create table6v2, include = TRUE}

caption_table <- "Table 6: Likelihood test between Model 5 and Model 6, which adds day as a random slope. Model 6 has significantly lower AIC and BIC, and a very small p-value, so it is statistically significant. Note that models are run using REML, as the have the same fixed effects.."


col_labels <- c("Model Version", 
                "\\shortstack{Degrees of\\\\Freedom}", 
                "AIC", 
                "BIC", 
                "Log-Likelihood", 
                "\\shortstack{Likelihood Ratio\\\\Test Stat}", 
                "P-Value")

kable(anova_mdl6_clean, 
      format="latex",
      col.names = col_labels,
      align = "lrrrrrr",
      booktabs = TRUE,#, #right aligns numbers
      escape = FALSE) %>% 
    #   caption = caption_table) %>%
    kable_styling(latex_options = "scale_down", 
                full_width = FALSE) %>%
    add_header_above(c("Table 6: Likelihood Ratio Test for Random Slope for day" = 7), 
                   bold = TRUE, align = "l") %>%
    row_spec(0, bold = TRUE) %>% #highlight main row
    footnote(
           # The caption will wrap inside this box and NOT shrink the table.
           general = paste0("\\\\begin{minipage}{20cm}\\\\vspace{2mm}\\\\footnotesize{", caption_table, "}\\\\end{minipage}"),
           general_title = "", 
           threeparttable = TRUE, 
           escape = FALSE)
```



## Section 5.7: Covariance Structures

The selected hierarchical structure (Model 6) effectively partitions the data's variance into three distinct components: global participant-level baselines, individual thermal adaptation patterns (represented by the random slope for day), and nested sensor-level deviations. This framework accounts for the inherent dependencies in the data while acknowledging that heat flow sensitivity is not uniform across the study population. 

However, this still imposes a strict assumption of conditional independence for the remaining residual errors. It has a compound symmetry structure, assuming a constant correlation structure where the correlation between Session 1 and Session 2 is identical to the correlation between Session 1 and Session 4, ignoring the fact that observations closer in time are typically more highly correlated. As such, we examined implementing an autoregressive (AR1) correlation structure for the residuals in Model 7. We assumed this was nested within person, sensor, and day, ensuring independence between people, sensor location, and days. We did not assume independence between activity because the activities were done in order, with session 1 consisting of sitting, then standing, then walking, then a break, and then session two beginning with the same cycle (although sessions 2 and 4 did not have walking). This assumes that these measurements were equally spaced in time, which is likely not true - timestamp data could improve this assumption.

The correlation structure has a $\phi$ of 0.5745, meaning that the correlation between sesson 1 and 2 is ~0.57, and session 1 & 3 is ~$0.57^2$. This is fairly high, suggesting the previous data point strongly influences the next one. Since we have the data segmented by session and activity, it suggests that sitting in session 1 impacts standing in session 1, and walking in session 1 impacts sitting in session 2. Additionally, Table 7 shows that Model 7 has significantly lower AIC and BIC, and is statistically significant in the likelihood ratio test against Model 6, suggesting an AR1 covariance structure is a significant improvement on the Compound Symmetry structure, and that there is a strong temporal auto-correlation.


```{r AR1 model}
mdl7_reml <- lme(
  fixed = log(HF_value) ~ (day + session + activity + sex) * HF_type, 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  data = data_clean, 
  method = "REML"
)
print(summary(mdl7_reml))
```

```{r anova mdl7}


anova_mdl7 <- anova(mdl6_reml, mdl7_reml)
anova_mdl7_clean <- data.frame(
  Model = c("Model 6: Assumes Compound Symmetry", "Model 7: Assumes Auto-regressive AR(1)"),
  Df = as.numeric(anova_mdl7$df),
  AIC = round(as.numeric(anova_mdl7$AIC)),
  BIC = round(as.numeric(anova_mdl7$BIC)),
  LogLik = round(as.numeric(anova_mdl7$logLik)),
  LRT_Stat = c("--", round(anova_mdl7$L.Ratio[2], 2)),
  p_value = anova_mdl7$`p-value`[2])

print(anova_mdl7)
```

```{r create table 7 v2, include = TRUE}

caption_table <- "Table 7: Likelihood test between Model 6 and Model 7, testing an AR1 structure where magnitude of correlation depends on amount of time between observations (instead of flat correlation). Model 7 has significantly lower AIC and BIC and a very small p-value, so it is statistically significant. Note that models are run using REML, as they have the same fixed effects."


col_labels <- c("Model Version", 
                "\\shortstack{Degrees of\\\\Freedom}", 
                "AIC", 
                "BIC", 
                "Log-Likelihood", 
                "\\shortstack{Likelihood Ratio\\\\Test Stat}", 
                "P-Value")

kable(anova_mdl7_clean, 
      format="latex",
      col.names = col_labels,
      align = "lrrrrrr",
      booktabs = TRUE,#, #right aligns numbers
      escape = FALSE) %>% 
    #   caption = caption_table) %>%
    kable_styling(latex_options = "scale_down", 
                full_width = FALSE) %>%
    add_header_above(c("Table 7: Likelihood Ratio Test for Covariance Structures" = 7), 
                   bold = TRUE, align = "l") %>%
    row_spec(0, bold = TRUE) %>% #highlight main row
    footnote(
           # The caption will wrap inside this box and NOT shrink the table.
           general = paste0("\\\\begin{minipage}{20cm}\\\\vspace{2mm}\\\\footnotesize{", caption_table, "}\\\\end{minipage}"),
           general_title = "", 
           threeparttable = TRUE, 
           escape = FALSE)
```



```{r AR1 model v2}
mdl7b_reml <- lme(
  fixed = log(HF_value) ~ (day + session + activity + sex) * HF_type, 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day/activity), 
  data = data_clean, 
  method = "REML"
)
print(summary(mdl7b_reml))
```

```{r anova mdl7b}


anova_mdl7b <- anova(mdl7_reml, mdl7b_reml)
# anova_mdl7b_clean <- data.frame(
#   Model = c("Model 6: Assumes Compound Symmetry", "Model 7: Assumes Auto-regressive AR(1)"),
#   Df = as.numeric(anova_mdl7b$df),
#   AIC = round(as.numeric(anova_mdl7b$AIC)),
#   BIC = round(as.numeric(anova_mdl7b$BIC)),
#   LogLik = round(as.numeric(anova_mdl7b$logLik)),
#   LRT_Stat = c("--", round(anova_mdl7b$L.Ratio[2], 2)),
#   p_value = anova_mdl7$`p-value`[2])

print(anova_mdl7b)
```

```{r create table7, include = TRUE}

caption_table <- "Table 7: Likelihood test between Model 6 and Model 7, testing an AR1 structure where magnitude of correlation depends on amount of time between observations (instead of flat correlation). Model 7 has significantly lower AIC and BIC and a very small p-value, so it is statistically significant. Note that models are run using REML, as they have the same fixed effects."


col_labels <- c("Model Version", 
                "\\shortstack{Degrees of\\\\Freedom}", 
                "AIC", 
                "BIC", 
                "Log-Likelihood", 
                "\\shortstack{Likelihood Ratio\\\\Test Stat}", 
                "P-Value")

kable(anova_mdl7_clean, 
      format="latex",
      col.names = col_labels,
      align = "lrrrrrr",
      booktabs = TRUE,#, #right aligns numbers
      escape = FALSE) %>% 
    #   caption = caption_table) %>%
    kable_styling(latex_options = "scale_down", 
                full_width = FALSE) %>%
    add_header_above(c("Table 7: Likelihood Ratio Test for Covariance Structures" = 7), 
                   bold = TRUE, align = "l") %>%
    row_spec(0, bold = TRUE) %>% #highlight main row
    footnote(
           # The caption will wrap inside this box and NOT shrink the table.
           general = paste0("\\\\begin{minipage}{20cm}\\\\vspace{2mm}\\\\footnotesize{", caption_table, "}\\\\end{minipage}"),
           general_title = "", 
           threeparttable = TRUE, 
           escape = FALSE)
```

Model 7 (AR(1)) showed that there is a temporal aspect to the data, where consecutive sessions are more closely linked than further ones. However, AR(1) assumes that the amount of variance is identical for all sensors. However, we would expect sensors on different body parts to have different variance. As such, we tested a heterogeneous AR structure ARH(1) by combining Model 7's AR(1) structure with heterogeneous residual variances across sensors. 

Model 8's $\phi$ is 0.6269, higher than Model 7's (0.5745), suggesting the temporal aspect of the data was higher than previously suggested and was masked by the noise levels of the sensors. The estimates for the heterogenuous variances relative to sensor 16 (baseline, dominant hand) are: 1.27 (sensor 2, chest (side)), 0.72 (sensor 3, upper chest), 0.35 (sensor 5, upper arm), 0.95 (sensor 7, lower arm), and 0.97 (sensor 8, abdomen). This implies that sensors on the abdomen, lower arm, and dominant hand have similar noise. The sensor on the chest (side) is significantly noisier than those sensors, and the sensors on the upper chest and arm are significantly less noisy. There does not seem to be an obvious trend or reason for this; it may be helpful to consult experts in the field. Additionally, Table 8 shows that the ARH(1) structure (Model 8) has significantly lower AIC and BIC than Model 7's AR(1) structure, and is statistically significant in the likelihood ratio test.

```{r ARH(1)}
mdl8_reml <- lme(
  fixed = log(HF_value) ~ (day + session + activity + sex) * HF_type, 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "REML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)
summary(mdl8_reml)
```


```{r create table8, include = TRUE}
anova_mdl8 <- anova(mdl7_reml, mdl8_reml)
print(anova_mdl8)

anova_mdl8_clean <- data.frame(
  Model = c("Model 7: Assumes Auto-regressive Structure (AR(1))", "Model 8: Assumes Heterogenuous AR Structure (ARH(1))"),
  Df = as.numeric(anova_mdl8$df),
  AIC = round(as.numeric(anova_mdl8$AIC)),
  BIC = round(as.numeric(anova_mdl8$BIC)),
  LogLik = round(as.numeric(anova_mdl8$logLik)),
  LRT_Stat = c("--", round(anova_mdl8$L.Ratio[2], 2)),
  p_value = anova_mdl8$`p-value`[2])


caption_table <- "Table 8: Likelihood test between Model 7 and Model 8, testing an ARH1 structure that allows each sensor to have its own unique residual variance instead of assuming uniform residual noise. Model 8 has significantly lower AIC and BIC and a very small p-value, so it is statistically significant. Note that models are run using REML, as they have the same fixed effects."


col_labels <- c("Model Version", 
                "\\shortstack{Degrees of\\\\Freedom}", 
                "AIC", 
                "BIC", 
                "Log-Likelihood", 
                "\\shortstack{Likelihood Ratio\\\\Test Stat}", 
                "P-Value")

kable(anova_mdl7_clean, 
      format="latex",
      col.names = col_labels,
      align = "lrrrrrr",
      booktabs = TRUE,#, #right aligns numbers
      escape = FALSE) %>% 
    #   caption = caption_table) %>%
    kable_styling(latex_options = "scale_down", 
                full_width = FALSE) %>%
    add_header_above(c("Table 7: Likelihood Ratio Test for Covariance Structures" = 7), 
                   bold = TRUE, align = "l") %>%
    row_spec(0, bold = TRUE) %>% #highlight main row
    footnote(
           # The caption will wrap inside this box and NOT shrink the table.
           general = paste0("\\\\begin{minipage}{20cm}\\\\vspace{2mm}\\\\footnotesize{", caption_table, "}\\\\end{minipage}"),
           general_title = "", 
           threeparttable = TRUE, 
           escape = FALSE)
```


```{r ARH(1) v2}
mdl8b_reml <- lme(
  fixed = log(HF_value) ~ (day + session + activity + sex) * HF_type, 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day/activity), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "REML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)

anova(mdl8_reml, mdl8b_reml)
```

## Section 5.8: Pruning Fixed Effects 

The final model utilizes an ARH(1) covariance structure to account for the temporal aspect of the data between sequential activities and sessions, while allowing for unique levels of noise at each sensor. Additionally, it has random intercepts for individuals and nested sensors to account for sensor-specific baselines per individual, and a person-level random slope for day to capture individual level responses to temperature changes. The model has become quite complex, so we re-examine fixed effects to determine if there are any that can be removed.

```{r prune_v1}
mdl8_ml <- lme(
  fixed = log(HF_value) ~ (day + session + activity + sex) * HF_type, 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "ML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)
```

```{r 9a}

mdl9a_ml_no_interact_sensor_sex <- lme(
  fixed = log(HF_value) ~ (day + session + activity) * HF_type + sex, 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "ML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)

mdl9a2_ml_no_sex <- lme(
  fixed = log(HF_value) ~ (day + session + activity) * HF_type , 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "ML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)
```

```{r 9b}
mdl9b_ml_no_interact_sensor_day <- lme(
  fixed = log(HF_value) ~ (sex + session + activity) * HF_type + day, 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "ML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)

mdl9b2_ml_no_day <- lme(
  fixed = log(HF_value) ~ (sex + session + activity) * HF_type , 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "ML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)

```

```{r 9c}
mdl9c_ml_no_interact_sensor_activity <- lme(
  fixed = log(HF_value) ~ (day + session + sex) * HF_type + activity, 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "ML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)

mdl9c2_ml_no_activity <- lme(
  fixed = log(HF_value) ~ (day + session + sex) * HF_type , 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "ML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)
```

```{r 9d}
mdl9d_ml_no_interact_sensor_session <- lme(
  fixed = log(HF_value) ~ (day + activity + sex) * HF_type + session, 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "ML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)

mdl9d2_ml_no_session <- lme(
  fixed = log(HF_value) ~ (day + activity + sex) * HF_type , 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "ML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)

```


```{r}
anova(mdl8_ml, mdl9a_ml_no_interact_sensor_sex)
#not statistically significant, higher BIC and AIC if include --> remove
```

```{r}
anova(mdl8_ml, mdl9a2_ml_no_sex)
#not statistically significant, higher BIC and AIC if include --> remove

```

```{r}
anova(mdl8_ml, mdl9b_ml_no_interact_sensor_day)
#lower AIC and BIC if include, and p-val <0.0001 --> include
```

```{r}
anova(mdl8_ml, mdl9b2_ml_no_day)
#lower AIC and BIC if include, and p-val <0.0001 --> include
```

```{r}
anova(mdl8_ml, mdl9c_ml_no_interact_sensor_activity)
#lower AIC and BIC if include, and p-val <0.0001 --> include

```

```{r}
anova(mdl8_ml, mdl9c2_ml_no_activity)
#lower AIC and BIC if include, and p-val <0.0001 --> include

```

```{r}
anova(mdl8_ml, mdl9d_ml_no_interact_sensor_session)
#BIC worsens (increases) significant if include, but AIC decreases slightly. p-value is 0.0021, stat significant but not crazy. --> remove

```

```{r}
anova(mdl8_ml, mdl9d2_ml_no_session)
#BIC worsens (increases) significant if include, but AIC decreases slightly (2 points). p-value is 0.0034, stat significant but not crazy. --> remove

```

```{r}

anova_summary_manual_mdl8 <- bind_rows(
  get_baseline_metrics(mdl8_ml, "Model 8 (Full Model)"),
  get_anova_outputs(mdl8_ml, mdl9a_ml_no_interact_sensor_sex, "Drop Sensor and Sex Interaction"),
  get_anova_outputs(mdl8_ml, mdl9a2_ml_no_sex, "Drop Sex Fully"),
  get_anova_outputs(mdl8_ml, mdl9b_ml_no_interact_sensor_day, "Drop Sensor and Day Interaction"),
  get_anova_outputs(mdl8_ml, mdl9b2_ml_no_day, "Drop Day Fully"),
  get_anova_outputs(mdl8_ml, mdl9c_ml_no_interact_sensor_activity, "Drop Sensor and Activity Interaction"),
  get_anova_outputs(mdl8_ml, mdl9c2_ml_no_activity, "Drop Activity Fully"),
  get_anova_outputs(mdl8_ml, mdl9d_ml_no_interact_sensor_session, "Drop Sensor and Session Interaction"),
  get_anova_outputs(mdl8_ml, mdl9d2_ml_no_session, "Drop Session Fully"),
)

print(anova_summary_manual_mdl8)
```

```{r create table 9}

caption_table <- "Table 9: Model selection of variables via backwards elimination. Model 8 includes the four interaction terms previously noted. Each subsequent row represents a model where either only one interaction term was dropped from Model 8 (Drop X and Y Interaction), or the whole variable and interaction (Drop Sex Fully means dropped the variable sex, and the interaction term sex and sensor). The LRT Statistic and P-Value are derived from Likelihood Ratio Tests comparing the reduced models to Model 8. Note all models were fit using Maximum Likelihood (ML), to compare across models."


col_labels <- c("Model Version", 
                "\\shortstack{Degrees of\\\\Freedom}", 
                "AIC", 
                "BIC", 
                "Log-Likelihood", 
                "\\shortstack{Likelihood Ratio\\\\Test Stat}", 
                "P-Value")

```


```{r run table 9, include = TRUE}

kable(anova_summary_manual_mdl8, 
      format="latex",
      col.names = col_labels,
      align = "lrrrrrr",
      booktabs = TRUE,#, #right aligns numbers
      escape = FALSE) %>% 
    #   caption = caption_table) %>%
    kable_styling(latex_options = "scale_down", 
                full_width = FALSE) %>%
    add_header_above(c("Table 9: Backwards Selection of Interaction Terms and Variables: Likelihood Ratio Test Summary" = 7), 
                   bold = TRUE, align = "l") %>%
    row_spec(0, bold = TRUE) %>% #highlight main row
    row_spec(1, italic = TRUE) %>%
    footnote(
           # This creates a box exactly the width of a standard page (15cm)
           # The caption will wrap inside this box and NOT shrink the table.
           general = paste0("\\\\begin{minipage}{15cm}\\\\vspace{2mm}\\\\footnotesize{", caption_table, "}\\\\end{minipage}"),
           general_title = "", 
           threeparttable = TRUE, 
           escape = FALSE)
```


This suggests that we should remove the variables Sex and the interaction term of Sensor and Sex, as the model without those terms has better AIC and BIC, and the p-value from the likelihood ratio test is not statistically significant (>0.4). We also removed the variable Session and the interaction term of Sensor and Session. Although they minimall improve the AIC (by <10 points), they worsen the BIC significantly and introduce ~18 degrees of freedom. 

This results in the final Model 9:  ($\log(\text{HF\_value}) \sim \text{activity + day + sensor +  sensor * activity + sensor * day}$), which evaluates the fixed effects of activity, day, sensor, and the interactions of sensor with activity and day. It addresses the hierarchical nature of the data by incorporating a random intercept for person, a nested random intercept for sensor, and a person-level random slope for day. Finally, the error structure is defined by an ARH(1) covariance structure, allowing temporal auto-correlation segmented by person, sensor, and day, and heterogenous residual variances to account for different levels of volatility across sensors.

```{r run model 9 final results, include = TRUE}
mdl_9_fin <- lme(
  fixed = log(HF_value) ~ (day + activity + sex) * HF_type , 
  random = list(
    person = ~ 1 + day,  
    HF_type = ~ 1        
  ), 
  correlation = corAR1(form = ~ 1 | person/HF_type/day), 
  weights = varIdent(form = ~ 1 | HF_type),
  data = data_clean, 
  method = "REML",
  control = lmeControl(msMaxIter = 200) #basically, increase iterations
)

summary(mdl_9_fin)
```


## Section 6: Discussion
