---
title: "Thermal Heat Flow Analysis"
author: "Nikita Gupta"
date: "March 13, 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup}
#, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
# echo = FALSE hides the code blocks in the final PDF
# warning/message = FALSE keeps the document clean of R's startup text
```

## Section 2: Examine and Edit dataset
```{r}
#, include=FALSE}
library(tidyverse) #installed tidyverse
library(nlme)
library(ggplot2)
library(here)
library(kableExtra)
library(modelsummary)

# Load the dataset
file_path <- here("analysis_and_plots/heat_flow_ICE_EPFL.csv")
df <- read.csv(file_path)

df_clean_v1 <- df %>% filter(!is.na(HF_value)) #remove rows with NA in HF_value
print(dim(df_clean_v1)) #2638 rows after removing NAs
print(dim(df)) #2736 rows 

# factorize dataset
df_clean <- df_clean_v1 %>%
  mutate(across(
    c(person, day, session, activity, sex, HF_type),
    as.factor
  ))
```


Now do plots to see trends

```{r}
# Summarize by person, day, session, sensor
df_grid_summary <- df_clean %>%
  group_by(person, day, session, HF_type) %>%
  summarize(mean_HF = mean(HF_value), .groups = 'drop')

# Create Faceted Spider Plot
ggplot(df_grid_summary, aes(x = session, y = mean_HF, group = person, color = person)) +
  geom_line(alpha = 0.4) +
  geom_point(size = 1) + stat_summary(aes(group = 1), fun = mean, geom = "line", 
               color = "black", size = 1.2) +
  facet_wrap (day ~ HF_type) #, scales = "free_y") + # Scales="free_y" lets you see the slopes clearly for each body part
  theme_minimal() +
  labs(title = "Heat Flow Evolution Across Sessions",
       subtitle = "Rows: Day (1=Cold, 2=Warm) | Columns: Sensor Type",
       x = "Session Number",
       y = "Mean Heat Flow (W/m²)") +
  theme(legend.position = "none")
```

## Section 3: Examine Residuals for Linear, Determine transformations (if any)


Look at residuals spread. by session, activity, days. NOT BY sensor or sex!
```{r}
# Base model with all main effects, no interactions, no person random effect
lm_base <- lm(HF_value ~ factor(session) + factor(activity)+ factor(HF_type) + factor(day) + factor(sex), data = df_clean)
df_clean$resids <- residuals(lm_base)


df_pivot_day_narem <- df_clean %>%
  pivot_wider(
    id_cols = c(person, session, activity, HF_type),
    names_from = day,
    values_from = resids,
    names_prefix = "res_day"
  ) %>%
  # Remove cases where data is missing for one of the two days
  filter(!is.na(res_day1) & !is.na(res_day2))

ggplot(df_pivot_day_narem, aes(x = res_day1, y = res_day2, color = factor(person))) +
  geom_point(alpha = 0.7, size = 2) +
  # Dashed line y=x represents perfect correlation
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") + 
  facet_grid(session ~ activity) +
  labs(
    title = "Residual Correlation: Day 1 vs Day 2",
    subtitle = "Checking for independence across different temperature environments",
    x = "Residuals from Day 1 (24°C)",
    y = "Residuals from Day 2 (18°C)",
    color = "Person"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


Look at residuals spread. by day, activity, and session. Not by sensor or sex!
```{r}
df_pivot <- df_clean %>%
  filter(session %in% c(1, 4)) %>%
  pivot_wider(
    id_cols = c(person, day, activity, HF_type),
    names_from = session,
    values_from = resids,
    names_prefix = "res_s"
  )

# Plot residuals of Session 1 vs Session 4 , by sensor and activity, colored by person to check for independence of measures across sessions
ggplot(df_pivot, aes(x = res_s1, y = res_s4, color = factor(person))) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") + 
  facet_grid(day ~ activity) +
  labs(
    title = "Residual Correlation: Session 1 vs Session 4",
    subtitle = "By Sensor and Activity | Colors represent individual participants",
    x = "Residuals from Session 1",
    y = "Residuals from Session 4",
    color = "Person"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


## Section 4: Run model with fixed effects
1. Run the log model with all fixed effects

```{r}
mdl1_log <- gls(
  log(HF_value) ~ day + session + activity + HF_type + sex,
  data   = df_clean,
  method = "ML"
)
print(summary(mdl1_log),4)
```

2. Run anovas to compare it to models without one of the variables (backwards selection)
```{r}
# print("No day")
mdl1_log_no_day <- gls(
  log(HF_value) ~ session + (activity) +  (HF_type) +  (sex), 
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log, mdl1_log_no_day))

# print("No session")
mdl1_log_no_session <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex), 
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log, mdl1_log_no_session))

# print("No activity")
mdl1_log_no_activity <- gls(
  log(HF_value) ~  (day) +  (session) +  (HF_type) +  (sex), 
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log, mdl1_log_no_activity))

# print("No sensor")
mdl1_log_no_sensor <- gls(
  log(HF_value) ~  (day) +  (session) +  (activity) +  (sex), 
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log, mdl1_log_no_sensor))

# print("No sex")
mdl1_log_no_sex <- gls(
  log(HF_value) ~  (day) +  (session) +  (activity) +  (HF_type), 
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log, mdl1_log_no_sex))


#Option 1: Run the "Drop-One" analysis
anova_summary <- drop1(mdl1_log, test = "Chisq")
print(anova_summary)


#option 2: Combine the anovas

#helper function to get anova outputs
get_baseline_metrics <- function(orig, name) {
  data.frame(
    Model = name,
    Df = attr(logLik(orig), "df"),
    AIC = round(AIC(orig), 4),
    BIC = round(BIC(orig), 4),
    LogLik = round(as.numeric(logLik(orig)), 4),
    stringsAsFactors = FALSE
  )
}

get_anova_outputs <- function(orig, new, name) {
  res <- anova(orig, new)
  data.frame(
    Model = name,
    Df = attr(logLik(new), "df"),
    AIC = round(res$AIC[2],4),
    BIC = round(res$BIC[2],4),
    LogLik = round(res$logLik[2],4),
    LRT_Stat = round(res$L.Ratio[2],4),
    p_value = round(res$`p-value`[2],4)
  )
}

anova_summary_manual_mdl1 <- bind_rows(
  get_baseline_metrics(mdl1_log, "Full Model"),
  get_anova_outputs(mdl1_log, mdl1_log_no_day, "Drop Day"),
  get_anova_outputs(mdl1_log, mdl1_log_no_session, "Drop Session"),
  get_anova_outputs(mdl1_log, mdl1_log_no_activity, "Drop Activity"),
  get_anova_outputs(mdl1_log, mdl1_log_no_sensor, "Drop Sensor"),
  get_anova_outputs(mdl1_log, mdl1_log_no_sex, "Drop Sex")
)

print(anova_summary_manual_mdl1)
```

The above says to not include session, so we remove it! Then we compare the effects of removing the other variables to that model.

```{r}
# print("No session and no day")
mdl1_log_no_session_no_day <- gls(
  log(HF_value) ~   (activity) +  (HF_type) +  (sex), 
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log_no_session, mdl1_log_no_session_no_day))

# print("No session and no activity")
mdl1_log_no_session_no_activity <- gls(
  log(HF_value) ~  (day) +  (HF_type) +  (sex), 
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log_no_session, mdl1_log_no_session_no_activity))

# print("No session and no sensor")
mdl1_log_no_session_no_sensor <- gls(
  log(HF_value) ~  (day) +  (activity) +  (sex), 
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log_no_session, mdl1_log_no_session_no_sensor))


# print("No session and no sex")
mdl1_log_no_session_no_sex <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type), 
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log_no_session, mdl1_log_no_session_no_sex))

anova_summary_manual_mdl1_nosession <- bind_rows(
  get_baseline_metrics(mdl1_log_no_session, "Current Model (No Session)"),
  get_anova_outputs(mdl1_log_no_session, mdl1_log_no_session_no_day, "Drop Day"),
  get_anova_outputs(mdl1_log_no_session, mdl1_log_no_session_no_activity, "Drop Activity"),
  get_anova_outputs(mdl1_log_no_session, mdl1_log_no_session_no_sensor, "Drop Sensor"),
  get_anova_outputs(mdl1_log_no_session, mdl1_log_no_session_no_sex, "Drop Sex")
)

print(anova_summary_manual_mdl1_nosession)
```

## Takeaway: Given the above, we propose a fixed effect model of mdl1_log_no_session:  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex).
- Activity results in a drop in BIC, so we could consider removing it
- Sex barely improves BIC, so we could consider removing it
- For now, including - will re-examine at the end.

## Section 5.1: Run model with interaction terms, v1
Now try the ten possible interaction terms.

```{r}
#HF_type*day
#HF_type*activity
#HF_type*sex
#HF_type*session
#day*activity
#day*sex
#day*session
#activity*sex
#activity*session
#sex*session
mdl2_log_no_session_all_interactives <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type *session) + (day * activity) + (day * sex) + (day * session) + (activity * sex) + (sex * session), #(activity * session)  - df issue (too many vars)
  data   = df_clean,
  method = "ML"
)
print(summary(mdl2_log_no_session_all_interactives))

#now: try running just the model with fixed effects and this interactive effect
#same issue:gls fit singular, rank 18...
# mdl2_log_no_session_interact_activity_session <- gls(
#   log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex)+ (activity * session), # - df issue (too many vars)
#   data   = df_clean,
#   method = "ML"
# )
```


We find that :
- p-values are too high for all coefficients (over 0.05) for the following; note that 0.05 is not a strict cutoff, but a good starting point to determine which interaction effects to examine:
#sex*session
#day*sex
#day*activity

#not enough degrees freedom:
#activity*session

## Section 5.2: Run model with interaction terms

Now, try adding in each interactive term one by one (of the ones not automatically excluded). Test with an ANOVA.

```{r}

#sex*session
mdl2_log_no_session_interact_sex_session<- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (sex * session) ,
  data   = df_clean,
  method = "ML"
)

#day*sex
dl2_log_no_session_interact_sex_day<- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (sex * day) ,
  data   = df_clean,
  method = "ML"
)

#day*activity
dl2_log_no_session_interact_day_activity<- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (activity * day) ,
  data   = df_clean,
  method = "ML"
)


#HF_type*day
mdl2_log_no_session_interact_sensor_day <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) ,
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_day))

#HF_type*activity
mdl2_log_no_session_interact_sensor_activity <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * activity) ,
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_activity))

#HF_type*sex
mdl2_log_no_session_interact_sensor_sex <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * sex) ,
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_sex))


#HF_type*session
mdl2_log_no_session_interact_sensor_session <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * session) ,
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_session))


#activity*sex
mdl2_log_no_session_interact_activity_sex <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (activity * sex) ,
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log_no_session, mdl2_log_no_session_interact_activity_sex))

#sex*session
mdl2_log_no_session_interact_session_sex <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (session * sex) ,
  data   = df_clean,
  method = "ML"
)
# print(anova(mdl1_log_no_session, mdl2_log_no_session_interact_session_sex))

#day*session
mdl2_log_no_session_interact_session_day <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (session * day) ,
  data   = df_clean,
  method = "ML"
)


anova_summary_manual_mdl1_nosession_interact1 <- bind_rows(
  get_baseline_metrics(mdl1_log_no_session, "Current Model (No Session)"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_day, "Interact Sensor & Day"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_activity, "Interact Sensor & Activity"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_sex, "Interact Sensor & Sex"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_sensor_session, "Interact Sensor & Session"),
  get_anova_outputs(mdl1_log_no_session, dl2_log_no_session_interact_day_activity, "Interact Activity & Day"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_activity_sex, "Interact Activity & Sex"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_sex_session, "Interact Session & Sex"),
  get_anova_outputs(mdl1_log_no_session, dl2_log_no_session_interact_sex_day, "Interact Day & Sex"),
  get_anova_outputs(mdl1_log_no_session, mdl2_log_no_session_interact_session_day, "Interact Day & Session")
)

#day*session
#activity*session

print(anova_summary_manual_mdl1_nosession_interact1)
```

This suggests we should include the below. Both their AIC and BIC decreases, and they have significant p-values from the ANOVA test.

- HF_type * day

- HF_type * activity


We should potentially include the below. Although their AIC decreases and they have significant p-values from the ANOVA test, their BIC increased.

- HF_type * sex 

- HF_type * session

- Day * session


And not include the below, as both their AIC and BIC increase, and they do not have significnat p-values from the ANOVA test.

- activity * day

- activity * sex

- session * sex

- day * sex


And we do not include the following due to issues with degrees of freedom:

- activity * sessino



## Section 5.3

Now, try adding all give of the suggested interactive variables, and then compare that to only having four interactive variables
```{r}
#HF_type*day, HF_type * activity, HF_type * sex, HF_type * session, day * session
mdl3 <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type * session) + (day * session),
  data   = df_clean,
  method = "ML"
)

#remove hf_type * day
mdl3a <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) +(HF_type * activity) + (HF_type * sex) + (HF_type * session) + (day * session),
  data   = df_clean,
  method = "ML"
)

#remove hf_type * activity
mdl3b <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) + (HF_type * sex) + (HF_type * session) + (day * session),
  data   = df_clean,
  method = "ML"
)

#remove hf_type * sex
mdl3c <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) + (HF_type * activity) + (HF_type * session) + (day * session),
  data   = df_clean,
  method = "ML"
)

#remove hf_type * session
mdl3d <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type * session),
  data   = df_clean,
  method = "ML"
)

#remove day * session
mdl3e <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (day * session),
  data   = df_clean,
  method = "ML"
)

anova_summary_manual_mdl1_nosession_interact2 <- bind_rows(
  get_baseline_metrics(mdl3, "Current Model"),
  get_anova_outputs(mdl3, mdl3a, "Drop Sensor & Day"),
  get_anova_outputs(mdl3, mdl3b, "Drop Sensor & Activity"),
  get_anova_outputs(mdl3, mdl3c, "Drop Sensor & Sex"),
  get_anova_outputs(mdl3, mdl3d, "Drop Sensor & Session"),
  get_anova_outputs(mdl3, mdl3e, "Drop Day & Session")
)
print(anova_summary_manual_mdl1_nosession_interact2)
```


Now, we see a similar story. 

Both AIC and BIC decrease (and p-value is stat. sig) when we include the following:

- Sensor & Day

- Sensor & Activity

However, BIC increases when we include the following:

- Sensor & Sex

- Sensor & Session

- Day & Session


For now, we will take that to mean to include all interactive variables! Note that including sensor & Session + Day & Session adds session back to the model, in general. We will remove variables later, after looking at random intercepts.




## Section 6: Run model with random intercepts

## Section 6.1: Does adding the person random intercept do anything (with interaction variables)?

Now, I will test if there's inter-person variation.

```{r}
#final model before random effects
#  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) 
#                   + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type * session),

mdl4_no_random_with_interact_reml <- gls(
  log(HF_value) ~  (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type * session) + (day * session) + (session),
  data   = df_clean,
  method = "REML"
)

mdl4_random_person_with_interact_reml <- lme(
  fixed  = log(HF_value) ~ (day) +  (activity) +  (HF_type) +  (sex) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type * session) + (day * session) + (session),
  random = ~ 1 | person,
  data   = df_clean,
  method = "REML"
)

#anova to compare
print(anova(mdl4_no_random_with_interact_reml, mdl4_random_person_with_interact_reml))

#now see if the variance of the random effect is 0.
print(summary(mdl4_random_person_with_interact_reml))

#yes - it stdev = 0.0795, so diff levels. however there is still leftover noise (0.2819). ICC is .073 (0.0795^2/(0.0795+0.2819)^2) --> meaning that person effect only explains 7.3% of unexplained variance.

```

Yes, the adding person as the random intercept does decrease AIC

### Section 6.2: Does adding HF_type as random intercept help?
```{r}

#hat about comparing the HF_type intercept with interaction effects, to the baseline person random intercept with interaction effects?
mdl4_random_person_sensor_with_interact_reml <- lme(
  fixed  = log(HF_value) ~ (day) +  (activity) +  (HF_type) +  (sex) + (session) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type * session) + (day * session), 
  random = ~ 1 | person/HF_type,
  data   = df_clean,
  method = "REML"
)
print(anova(mdl4_random_person_sensor_with_interact_reml, mdl4_random_person_with_interact_reml))


#now see if the variance of the random effect is 0.
print(summary(mdl4_random_person_sensor_with_interact_reml))

#it is not! actually, sensor is more important than person.
```


## Section 6.5: Check other random intercepts.
```{r testname3}
mdl4_random_person_sensor_activity_with_interact_reml <- lme(
  fixed  = log(HF_value) ~ (day) +  (activity) +  (HF_type) +  (sex) + (session) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type * session)+ (day * session), 
  random = ~ 1 | person/HF_type/activity,
  data   = df_clean,
  method = "REML"
)

#anova to compare
print(anova(mdl4_random_person_sensor_with_interact_reml, mdl4_random_person_sensor_activity_with_interact_reml))

mdl4_random_person_sensor_day_with_interact_reml<- lme(
  fixed  = log(HF_value) ~ (day) +  (activity) +  (HF_type) +  (sex) + (session) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type * session)+ (day * session), 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "REML"
)

#how much of variance does this explain
print(summary(mdl4_random_person_sensor_day_with_interact_reml))
#day explains a lot of the unexplained residuals!! we should keep that.


#anova to compare
print(anova(mdl4_random_person_sensor_with_interact_reml, mdl4_random_person_sensor_day_with_interact_reml))

```

## Section 7: Let's test removing the variables we were initially unsure of (overlaps with high p-values from final model above)

Stuff with high pvalues:

- day

- sex

- session (although session 4 is significant)

- sensor & sex

- sensor & session (some)


```{r testname2}
mdl4_random_person_sensor_day_with_interact<- lme(
  fixed  = log(HF_value) ~ (day) +  (activity) +  (HF_type) +  (sex) + (session) + (HF_type * day) + (HF_type * activity) + (HF_type * sex) + (HF_type * session)+ (day * session), 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "ML"
)

#what if I remove the interactive variables that I was unsure of?
mdl4_random_person_sensor_day_with_interact_nodaysession <- lme(
  fixed  = log(HF_value) ~ (day) +  (activity) +  (HF_type) +  (session) + (sex) + (HF_type * day) +(HF_type * sex) + (HF_type * activity) + (HF_type * session) , 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "ML"
)

mdl4_random_person_sensor_day_with_interact_nosensorsession <- lme(
  fixed  = log(HF_value) ~ (day) +  (activity) +  (HF_type) +  (session) + (sex) + (HF_type * day) +(HF_type * sex) + (HF_type * activity) +  (day*session), 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "ML"
)

mdl4_random_person_sensor_day_with_interact_nosensorsex <- lme(
  fixed  = log(HF_value) ~ (day) +  (activity) +  (HF_type) +  (session) + (sex) + (HF_type * day)  + (HF_type * activity) + (HF_type * session)  +  (day*session), 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "ML"
)

#remove sex (and thus no sensor sex)
mdl4_random_person_sensor_day_with_interact_nosex <- lme(
  fixed  = log(HF_value) ~ (day) +  (activity) +  (HF_type) +  (session) + (HF_type * day)  + (HF_type * activity) + (HF_type * session)  +  (day*session), 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "ML"
)

#remove day (and htus interactions)
mdl4_random_person_sensor_day_with_interact_noday <- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (session) + (HF_type * activity) + (HF_type * session) + (sex) + (HF_type * sex), 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "ML"
)

mdl4_random_person_sensor_day_with_interact_nosession<- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (HF_type * activity) + (HF_type * day) + (sex) + (HF_type * sex), 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "ML"
)

mdl5_activity_sensor_day_session_sensoractivity_sensorday_sessionday<- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (session) + (HF_type * activity) + (HF_type * day) + (day * session), 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "ML"
)

mdl5_activity_sensor_day_sensoractivity_sensorday<- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day)  + (HF_type * activity) + (HF_type * day), 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "ML"
)

anova_summary_manual_mdl4_nosession_interact <- bind_rows(
  get_baseline_metrics(mdl4_random_person_sensor_day_with_interact, "Current Model"),
  get_anova_outputs(mdl4_random_person_sensor_day_with_interact, mdl4_random_person_sensor_day_with_interact_nodaysession, "Drop Day & Session"),
  get_anova_outputs(mdl4_random_person_sensor_day_with_interact, mdl4_random_person_sensor_day_with_interact_nosensorsession, "Drop Sensor & Session"),
  get_anova_outputs(mdl4_random_person_sensor_day_with_interact, mdl4_random_person_sensor_day_with_interact_nosensorsex, "Drop Sensor & Sex"),
  get_anova_outputs(mdl4_random_person_sensor_day_with_interact, mdl4_random_person_sensor_day_with_interact_nosex, "Drop Sex"),
  get_anova_outputs(mdl4_random_person_sensor_day_with_interact, mdl4_random_person_sensor_day_with_interact_noday, "Drop Day"),
  get_anova_outputs(mdl4_random_person_sensor_day_with_interact, mdl4_random_person_sensor_day_with_interact_nosession, "Drop Session"),
  get_anova_outputs(mdl4_random_person_sensor_day_with_interact, mdl5_activity_sensor_day_session_sensoractivity_sensorday_sessionday, "Final"),
  get_anova_outputs(mdl4_random_person_sensor_day_with_interact, mdl5_activity_sensor_day_sensoractivity_sensorday, "Final no session")
)
print(anova_summary_manual_mdl4_nosession_interact)

```


We should definitely drop sensor and sex. 

Dropping day & session does nothing - AIC+BIC increase, and p-value is 0 for it. So we will keep it.

Dropping sensor & session increases AIC & decreases BIC, so we will drop it.

Just dropping sensor & session is better than dropping session overall, so we will keep session.

Dropping sensor & sex is unequivocally good. Relatedly, dropping sex overall is even better. So we will drop everything sex.

Dropping day overall is not good, so we will keep it.

Propose to drop sex, sensor & sex, sensor & session.

## Takeaway: Final model: fixed = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (session) + (HF_type * activity) + (HF_type * day) + (day * session), , random = ~ 1 | person/HF_type/day,



## Section 8: Look at random slopes!

Initial hypothesis: people cool down at different rates. Try adding random slope for session. However, there's convergence issues...

```{r testname}
#hypothesis: people cool down at different rates


mdl5_activity_sensor_day_session_sensoractivity_sensorday_sessionday_reml<- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (session) + (HF_type * activity) + (HF_type * day) + (day * session), 
  random = ~ 1 | person/HF_type/day,
  data   = df_clean,
  method = "REML"
)
print(summary(mdl5_activity_sensor_day_session_sensoractivity_sensorday_sessionday_reml))
```

```{r testname123}
# mdl6_rs_session_reml<- lme(
#   fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (session) + (HF_type * activity) + (HF_type * day) + (day * session), 
#   random = list(person = ~ session, HF_type = ~ 1, day = ~ 1),
#   data   = df_clean,
#   method = "REML",
#   control = lmeControl(maxIter = 200, 
#                          msMaxIter = 200, 
#                          tolerance = 1e-6, 
#                          niterEM = 100,
#                          opt = "nlminb") #basically, increase iterations
# )

# print(anova(mdl5_activity_sensor_day_session_sensoractivity_sensorday_sessionday_reml, mdl6_rs_session_reml))

# print(summary(mdl6_rs_session_reml))

#for random effects, we care about the variance! --> is the stdev high?
# for fixed effects, just the estimates --> anova, AIC/BIC, pval
```

It fails when we try to do this for REML. Instead, try numerical. (Also works because when ran for ML, correlation between the sessions was high)

```{r test_session_2}
df_clean$session_num <- as.numeric(as.character(df_clean$session))

mdl6_rs_session_num_reml<- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (session) + (HF_type * activity) + (HF_type * day) + (day * session), 
  random = list(person = ~ session_num, HF_type = ~ 1, day = ~ 1),
  data   = df_clean,
  method = "REML",
  control = lmeControl(maxIter = 200, 
                         msMaxIter = 200, 
                         tolerance = 1e-6, 
                         niterEM = 100,
                         opt = "nlminb") #basically, increase iterations
)

print(anova(mdl5_activity_sensor_day_session_sensoractivity_sensorday_sessionday_reml, mdl6_rs_session_num_reml))
print(summary(mdl6_rs_session_num_reml))

```


Ok now make everything numerical. 

```{r anothertest}
mdl6_rs_session_num_full_reml <- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (session_num) + (HF_type * activity) + (HF_type * day) + (day * session_num), 
  random = list(person = ~ session_num, HF_type = ~ 1, day = ~ 1),
  data   = df_clean,
  method = "REML",
  control = lmeControl(maxIter = 200, 
                         msMaxIter = 200, 
                         tolerance = 1e-6, 
                         niterEM = 100,
                         opt = "nlminb") #basically, increase iterations
)

print(anova(mdl6_rs_session_num_reml, mdl6_rs_session_num_full_reml))

print(summary(mdl6_rs_session_num_full_reml))

```




## Section 9: Look at different covariance structures
Normal residuals
```{r }
mdl6_rs_session_num_res <- residuals(mdl6_rs_session_num_reml, type = "normalized")
acf(mdl6_rs_session_num_res, main = "ACF of BAU residuals")
```

```{r}
# Base model with all main effects, no interactions, no person random effect
df_clean$resids_fin <- residuals(mdl6_rs_session_num_reml)


df_pivot <- df_clean %>%
  pivot_wider(
    id_cols = c(person, session, activity, HF_type),
    names_from = day,
    values_from = resids_fin,
    names_prefix = "res_day"
  ) %>%
  # Remove cases where data is missing for one of the two days
  filter(!is.na(res_day1) & !is.na(res_day2))

ggplot(df_pivot, aes(x = res_day1, y = res_day2, color = factor(person))) +
  geom_point(alpha = 0.7, size = 2) +
  # Dashed line y=x represents perfect correlation
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") + 
  facet_grid(session ~ activity) +
  labs(
    title = "Residual Correlation: Day 1 vs Day 2 (Independent)",
    subtitle = "Checking for independence across different temperature environments",
    x = "Residuals from Day 1 (24°C)",
    y = "Residuals from Day 2 (18°C)",
    color = "Person"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r 123}
mdl6_rs_session_num_ar1temp_reml<- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (session) + (HF_type * activity) + (HF_type * day) + (day * session), 
  random = list(person = ~ session_num, HF_type = ~ 1, day = ~ 1),
  data   = df_clean,
  method = "REML",
  correlation = corAR1(form = ~ 1 | person/HF_type/day),
  control = lmeControl(maxIter = 200, 
                         msMaxIter = 200, 
                         tolerance = 1e-6, 
                         niterEM = 100,
                         opt = "nlminb") #basically, increase iterations
)
```

```{r 345}
mdl6_rs_session_num_ar1temp_res <- residuals(mdl6_rs_session_num_ar1temp_reml, type = "normalized",)

```

```{r 1254}
acf(mdl6_rs_session_num_ar1temp_res, main = "ACF of AR1 residuals")
```

```{r 164}

print(anova(mdl6_rs_session_num_reml, mdl6_rs_session_num_ar1temp_reml))
```


```{r}
# Base model with all main effects, no interactions, no person random effect
df_clean$resids_fin_ar1 <- residuals(mdl6_rs_session_num_ar1temp_reml)


df_pivot <- df_clean %>%
  pivot_wider(
    id_cols = c(person, session, activity, HF_type),
    names_from = day,
    values_from = resids_fin_ar1,
    names_prefix = "res_day"
  ) %>%
  # Remove cases where data is missing for one of the two days
  filter(!is.na(res_day1) & !is.na(res_day2))

ggplot(df_pivot, aes(x = res_day1, y = res_day2, color = factor(person))) +
  geom_point(alpha = 0.7, size = 2) +
  # Dashed line y=x represents perfect correlation
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") + 
  facet_grid(session ~ activity) +
  labs(
    title = "Residual Correlation: Day 1 vs Day 2 (AR1)",
    subtitle = "Checking for independence across different temperature environments",
    x = "Residuals from Day 1 (24°C)",
    y = "Residuals from Day 2 (18°C)",
    color = "Person"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```



```{r }
mdl6_rs_session_num_arh1temp_reml_day<- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (session) + (HF_type * activity) + (HF_type * day) + (day * session), 
  random = list(person = ~ session_num, HF_type = ~ 1, day = ~ 1),
  data   = df_clean,
  method = "REML",
  correlation = corAR1(form = ~ 1 | person/HF_type/day),
  weights = varIdent(form = ~ 1 | day),
  control = lmeControl(maxIter = 200, 
                         msMaxIter = 200, 
                         tolerance = 1e-6, 
                         niterEM = 100,
                         opt = "nlminb") #basically, increase iterations
)

print(anova(mdl6_rs_session_num_ar1temp_reml, mdl6_rs_session_num_arh1temp_reml_day))


mdl6_rs_session_num_arh1temp_reml_day_res <- residuals(mdl6_rs_session_num_arh1temp_reml_day, type = "normalized")
acf(mdl6_rs_session_num_arh1temp_reml_day_res, main = "ACF of ARH1 residuals (day)")
```

```{r}
# Base model with all main effects, no interactions, no person random effect
df_clean$resids_fin_arh1_day <- residuals(mdl6_rs_session_num_arh1temp_reml_day)


df_pivot <- df_clean %>%
  pivot_wider(
    id_cols = c(person, session, activity, HF_type),
    names_from = day,
    values_from = resids_fin_arh1_day,
    names_prefix = "res_day"
  ) %>%
  # Remove cases where data is missing for one of the two days
  filter(!is.na(res_day1) & !is.na(res_day2))

ggplot(df_pivot, aes(x = res_day1, y = res_day2, color = factor(person))) +
  geom_point(alpha = 0.7, size = 2) +
  # Dashed line y=x represents perfect correlation
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") + 
  facet_grid(session ~ activity) +
  labs(
    title = "Residual Correlation: Day 1 vs Day 2 (ARH1, Day)",
    subtitle = "Checking for independence across different temperature environments",
    x = "Residuals from Day 1 (24°C)",
    y = "Residuals from Day 2 (18°C)",
    color = "Person"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```




```{r }
mdl6_rs_session_num_arh1temp_reml_activity<- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (session) + (HF_type * activity) + (HF_type * day) + (day * session), 
  random = list(person = ~ session_num, HF_type = ~ 1, day = ~ 1),
  data   = df_clean,
  method = "REML",
  correlation = corAR1(form = ~ 1 | person/HF_type/day),
  weights = varIdent(form = ~ 1 | activity),
  control = lmeControl(maxIter = 200, 
                         msMaxIter = 200, 
                         tolerance = 1e-6, 
                         niterEM = 100,
                         opt = "nlminb") #basically, increase iterations
)

print(anova(mdl6_rs_session_num_ar1temp_reml, mdl6_rs_session_num_arh1temp_reml_activity))


mdl6_rs_session_num_arh1temp_reml_activity_res <- residuals(mdl6_rs_session_num_arh1temp_reml_activity, type = "normalized")
acf(mdl6_rs_session_num_arh1temp_reml_activity_res, main = "ACF of ARH1 residuals (actiity)")
```

```{r}
# Base model with all main effects, no interactions, no person random effect
df_clean$resids_fin_arh1_activity <- residuals(mdl6_rs_session_num_arh1temp_reml_activity)


df_pivot <- df_clean %>%
  pivot_wider(
    id_cols = c(person, session, activity, HF_type),
    names_from = day,
    values_from = resids_fin_arh1_activity,
    names_prefix = "res_day"
  ) %>%
  # Remove cases where data is missing for one of the two days
  filter(!is.na(res_day1) & !is.na(res_day2))

ggplot(df_pivot, aes(x = res_day1, y = res_day2, color = factor(person))) +
  geom_point(alpha = 0.7, size = 2) +
  # Dashed line y=x represents perfect correlation
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") + 
  facet_grid(session ~ activity) +
  labs(
    title = "Residual Correlation: Day 1 vs Day 2 (ARH1, Activity)",
    subtitle = "Checking for independence across different temperature environments",
    x = "Residuals from Day 1 (24°C)",
    y = "Residuals from Day 2 (18°C)",
    color = "Person"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```


```{r }
mdl6_rs_session_num_arh1temp_reml_sensor<- lme(
  fixed  = log(HF_value) ~  (activity) +  (HF_type) +  (day) + (session) + (HF_type * activity) + (HF_type * day) + (day * session), 
  random = list(person = ~ session_num, HF_type = ~ 1, day = ~ 1),
  data   = df_clean,
  method = "REML",
  correlation = corAR1(form = ~ 1 | person/HF_type/day),
  weights = varIdent(form = ~ 1 | HF_type),
  control = lmeControl(maxIter = 200, 
                         msMaxIter = 200, 
                         tolerance = 1e-6, 
                         niterEM = 100,
                         opt = "nlminb") #basically, increase iterations
)

print(anova(mdl6_rs_session_num_ar1temp_reml, mdl6_rs_session_num_arh1temp_reml_sensor))


mdl6_rs_session_num_arh1temp_reml_sensor_res <- residuals(mdl6_rs_session_num_arh1temp_reml_sensor, type = "normalized")
acf(mdl6_rs_session_num_arh1temp_reml_sensor_res, main = "ACF of ARH1 residuals (sensor)")
```

```{r}
# Base model with all main effects, no interactions, no person random effect
df_clean$resids_fin_arh1_sensor <- residuals(mdl6_rs_session_num_arh1temp_reml_sensor)


df_pivot <- df_clean %>%
  pivot_wider(
    id_cols = c(person, session, activity, HF_type),
    names_from = day,
    values_from = resids_fin_arh1_sensor,
    names_prefix = "res_day"
  ) %>%
  # Remove cases where data is missing for one of the two days
  filter(!is.na(res_day1) & !is.na(res_day2))

ggplot(df_pivot, aes(x = res_day1, y = res_day2, color = factor(person))) +
  geom_point(alpha = 0.7, size = 2) +
  # Dashed line y=x represents perfect correlation
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") + 
  facet_grid(session ~ activity) +
  labs(
    title = "Residual Correlation: Day 1 vs Day 2 (ARH1, Sensor)",
    subtitle = "Checking for independence across different temperature environments",
    x = "Residuals from Day 1 (24°C)",
    y = "Residuals from Day 2 (18°C)",
    color = "Person"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

Final: do arh1 for sensor??